---
title: "Birds_National"
author: "Ron Chen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
#knitr::opts_chunk$set(dev = "cairo_pdf", dev.args=list(cairo_pdf = list(family='Arial')))

# pcks <- list('devtools','rlang','ggplot2','data.table', 'dplyr','mvabund','kableExtra','lme4','MASS','vegan',
#              'jtools', 'interactions','car','ecoCopula','emmeans','Cairo','extrafont', 'gllvm')
# sapply(pcks, require, character = TRUE)
library (devtools)
library (rlang)
library (ggplot2)
library (data.table)
library (dplyr)
library (mvabund)
library (kableExtra)
library (lme4)
library (MASS)
library (vegan)
library (jtools)
library (interactions)
library (car)
library (ecoCopula)
library (emmeans)
library (Cairo)
library (extrafont)

source("plot_model_effect.R")
source("prepare_bird_data_with_Z_filter.R")
source("plot_alpha_diversity.R")
source("plot_coefs_with_traits.R")
source("plot_coefs_with_traits_for_publication.R")
source("multiplot.R")
source("plot_individual_species_national.R")

rare_min_samp_thr_national <- 20
rare_min_samp_thr_national.all_data <- 80
rare_min_units_thr_national.all_data <- 3
export_all_plots <- FALSE
export_first_plot <- TRUE

# define font name
SoN_fontname <- "Almoni ML v5 AAA"

# Define the output folder path
output_path <- file.path("..","output","national")

# Check if the output folder exists, and create it if it doesn't
if (!dir.exists(output_path)) {
  dir.create(output_path, recursive = TRUE)
  message("Folder '", output_path, "' has been created.")
} else {
  message("Folder '", output_path, "' already exists.")
}
```

```{r load using function}
out <- prepare_bird_data()
var_names <- names(out)
for (i in 1:length(var_names)) {
  assign(var_names[i], out[[i]])
}
rm(out)
P <- P_byplot
P_no_rare <- P_byplot_no_rare
```

## National analysis - analyze ALL data from all units

```{r}
# define rare species
spec_names_no_rare_by_samples <- which(inc_by_spp_no_rare[,dplyr::select(.SD, any_of(unique(Araw[,SciName])))] %>% colSums() >= rare_min_samp_thr_national.all_data) %>% names()
tmp99 <- inc_by_spp_no_rare[,dplyr::select(.SD, any_of(c("unit",unique(Araw[,SciName]))))] %>%
    melt.data.table(variable.name = "SciName", value.name = "is.present", id.vars = "unit")
spec_names_no_rare_by_units <- tmp99[is.present==TRUE, .(num_units=uniqueN(unit)), by=SciName][num_units>=rare_min_units_thr_national.all_data,SciName]
spec_names_no_rare <- intersect(spec_names_no_rare_by_samples, spec_names_no_rare_by_units)
```

**Species in NATIONAL analysis, all plots, minimum number of plots `r rare_min_samp_thr_national.all_data` and minimum number of units `r rare_min_units_thr_national.all_data``:**

```{r}
unique(A[SciName %in% spec_names_no_rare, .(SciName, HebName)])[order(SciName)][,.(.I, SciName, HebName)] %>% kbl %>%  kable_styling(font_size = 11)
```

```{r}
unique(abu_by_spp_no_rare[,.(num_samples=uniqueN(point_name)), keyby=.(unit, subunit, agriculture, settlements, land_use, habitat, dunes, year)]) %>% kbl %>%  kable_styling(font_size = 11)

spp_no_rare <- abu_by_spp_no_rare[,dplyr::select(.SD, all_of(spec_names_no_rare))]
# remove sites where none of the species were observed for the entire monitoring period
samps2remove <- which(rowSums(spp_no_rare)==0)
```

**The following plots had zero total abundance:**

```{r}
abu_by_spp_no_rare[samps2remove, .(num_years=uniqueN(year)), keyby = point_name] %>% kbl %>%  kable_styling(font_size = 11)
```

**Only three plots had zeros for more than one year - no plots or sites to be removed.**

```{r}
# spp_no_rare <- mvabund(spp_no_rare[!samps2remove])
# env_data <- abu_by_spp_no_rare[!samps2remove,..col_names]
spp_no_rare <- mvabund(spp_no_rare)
env_data <- abu_by_spp_no_rare[,..col_names]
factor_col_names <- c("unit","subunit","site","settlements","agriculture","habitat","dunes","land_use","point_name")
env_data[,lapply(.SD,FUN = is.factor), .SDcols = factor_col_names] # already factors, but need to eliminate unnecessary levels
env_data[,(factor_col_names):=lapply(.SD,as.factor), .SDcols = factor_col_names]

# create campaign variable
env_data[,campaign:=ceiling((year-2014)/2)][year==2012,campaign:=0]

# create unique site variable
env_data[,unit_site:=as.factor(paste(unit,site,sep = "_"))]

print("abundance observations, aggregated into plots and species (i.e., several observations from the same species in the same plot are aggregated):")
plot(sort(spp_no_rare[spp_no_rare>0]))
print("zoom in on high abundance observations:")
dotchart(sort(A_no_rare[count_under_250>=30,count_under_250]))
meanvar.plot(spp_no_rare, xlab = "mean abundance of a given species across sites", ylab = "variance of the abundance of a given species across sites")
```

explore variables


```{r}
cat("\n\nTotal number of samples:\n\n",nrow(env_data))
kable(summary(env_data)) %>% kable_styling(font_size = 11)
pairs(env_data[,lapply(X = .SD,FUN = as.numeric), .SDcols = c("unit","site","unit_site","year","campaign","pilot","cos_td_rad","sin_td_rad")])
kable(cor(env_data[,lapply(X = .SD,FUN = as.numeric), .SDcols = c("unit","site","unit_site","year","campaign","pilot","cos_td_rad","sin_td_rad")], use = "pairwise.complete.obs")) %>% kable_styling(font_size = 11)
```

start model specification:

```{r community analysis MVabund model specification}
mva_m0.po <- manyglm(formula = spp_no_rare ~ year_ct + unit * cos_td_rad + unit * sin_td_rad, family = "poisson", data = env_data)
mva_m0.nb <- manyglm(formula = spp_no_rare ~ year_ct + unit * cos_td_rad + unit * sin_td_rad, family = "negative.binomial", data = env_data)
aic_dt <- data.table(nb = mva_m0.nb$aic, po=mva_m0.po$aic)
colMeans(aic_dt)
print("POISSON")
plot(mva_m0.po, which=1:3)
print("NEGATIVE BINOMIAL")
plot(mva_m0.nb, which=1:3)
```

negative binomial model is better than poisson according to residuals and AIC comparison.

```{r community analysis MVabund model specification 2}
mva_m0 <- mva_m0.nb
# mva_m1 <- manyglm(formula = spp_no_rare ~ year_ct + unit_site*cos_td_rad + unit_site*sin_td_rad + site, data = env_data)
# aic_dt$nb1 <- mva_m1$aic
mva_m2 <- manyglm(formula = spp_no_rare ~ year_ct + unit_site + cos_td_rad + sin_td_rad, data = env_data)
aic_dt$nb2 <- mva_m2$aic
mva_m3 <- manyglm(formula = spp_no_rare ~ year_ct + unit + cos_td_rad + sin_td_rad, data = env_data)
aic_dt$nb3 <- mva_m3$aic
colMeans(aic_dt)
```

The addition of the explanatory variable 'site' is somewhat improving the AIC of the model. Start with unit and site.
stepwise selection of model:

```{r community analysis MVabund model selection}
mva_m1 <- mva_m2
drop1(mva_m1)
```

remove cosine, probably not significant for all species

```{r}
mva_m1 <- manyglm(formula = spp_no_rare ~ year_ct + unit_site + sin_td_rad, data = env_data)
drop1(mva_m1)
```

final model year, unit_site and sampling time of year.

```{r}
####model validation####
plot(mva_m1, which=1:3)
```

```{r model interpretation - anova}
####model interpretation####
# summ_m1 <- summary(mva_m1, nBoot=99)
# saveRDS(summ_m1, file = "../output/National_MVabund_summary_output_99iter_min_samp80.rds")
# anov_m1.uni <- anova(mva_m1, p.uni = "adjusted", nBoot = 99, show.time = "all")
# saveRDS(anov_m1.uni, file = "../output/National_MVabund_anova_output_99iter_min_samp80.rds")
summ_m1 <- readRDS(file = "../output/National_MVabund_summary_output_99iter_min_samp80.rds")
print(summ_m1)
anov_m1.uni <- readRDS(file = "../output/National_MVabund_anova_output_99iter_min_samp80.rds")
print(anov_m1.uni)
```

plot coefficients:

```{r}
####plot coefs####
coefp <- merge(data.table(t(coef(mva_m1)/log(2)),keep.rownames=TRUE), data.table(t(anov_m1.uni$uni.p), keep.rownames = TRUE), by = "rn") # coefficients are on log2 scale to simplify interpretation, yet keep presenetation on log scale rather than linear. The link function used for poisson is log -> above 0 is a positive trend and below 0 is negative trend

# add total species abundance
coefp <- merge(coefp,as.data.table(colSums(spp_no_rare),keep.rownames = TRUE), by.x = "rn", by.y = "V1")

# add hebrew name and traits
sci_heb <- unique(A[,.(SciName,HebName)])
sci_heb[,SciName:=make.names(SciName)]
coefp <- merge(coefp, sci_heb, by.x = "rn", by.y = "SciName")

# set column names based on the predictor variables
colnames(coefp)[grepl("^rn$",colnames(coefp))] <- "SciName"
colnames(coefp)[grepl("Intercept).x",colnames(coefp))] <- "intercept.coef"
colnames(coefp)[grepl("Intercept).y",colnames(coefp))] <- "intercept.p"
colnames(coefp)[colnames(coefp)=="settlementsNear"] <- "settlementsNear.coef"
colnames(coefp)[colnames(coefp)=="settlements"] <- "settlements.p"
colnames(coefp)[colnames(coefp)=="agricultureNear"] <- "agricultureNear.coef"
colnames(coefp)[colnames(coefp)=="agriculture"] <- "agriculture.p"
colnames(coefp)[grepl("cos_td_rad.x",colnames(coefp))] <- "cos_td_rad.coef"
colnames(coefp)[colnames(coefp)=="cos_td_rad.y"] <- "cos_td_rad.p"
colnames(coefp)[grepl("sin_td_rad.x",colnames(coefp))] <- "sin_td_rad.coef"
colnames(coefp)[grepl("sin_td_rad.y",colnames(coefp))] <- "sin_td_rad.p"
colnames(coefp)[colnames(coefp)=="year_ct.x"] <- "year_ct.coef"
colnames(coefp)[colnames(coefp)=="year_ct.y"] <- "year_ct.p"
colnames(coefp)[colnames(coefp)=="V2"] <- "species_abundance"
colnames(coefp)[colnames(coefp)=="dunesshifting"] <- "dunesShifting.coef"
colnames(coefp)[colnames(coefp)=="dunes"] <- "dunes.p"
colnames(coefp)[colnames(coefp)=="site"] <- "site.p"
colnames(coefp)[colnames(coefp)=="unit_site"] <- "unit_site.p"
colnames(coefp)[colnames(coefp)=="subunitCarmel"] <- "subunitCarmel.coef"
colnames(coefp)[colnames(coefp)=="subunitGalilee"] <- "subunitGalilee.coef"
colnames(coefp)[colnames(coefp)=="subunit"] <- "subunit.p"
colnames(coefp)[colnames(coefp)=="habitatWadi"] <- "habitatWadi.coef"
colnames(coefp)[colnames(coefp)=="habitat"] <- "habitat.p"

plot_coefs_year <- plot_coefs_with_traits_for_publication(D = A_no_rare,
                                                          cp = coefp[,.(SciName,year_ct.coef,year_ct.p,species_abundance)],
                                                          plot_title = "Temporal effect on species abundance",
                                                          plot_xlabel = expression(paste("coefficient of year counter, (log"[2]," scale)")),
                                                          mark_batha = TRUE,
                                                          show_legend = FALSE,
                                                          export_plot = export_first_plot,
                                                          fontname = SoN_fontname,
                                                          effect_str = "year_ct",
                                                          outpath = output_path)
plot_coefs_year
```

## Plot individual species
### Streptopelia turtur

```{r, error=TRUE}
plot_individual_species_national(spec_name = "Streptopelia turtur", spec_var_name = "Streptopelia.turtur", model = mva_m1)
                                 # at_list = list(unit="Herbaceous and Dwarf-Shrub Vegetation", site="Natur"))
```

### Acridotheres tristis

```{r, error=TRUE}
plot_individual_species_national(spec_name = "Acridotheres tristis", spec_var_name = "Acridotheres.tristis",
                                 at_list = list(unit_site="Mediterranean-Desert Transition Zone_Beit Yatir"),
                                 model = mva_m1)
```

### Turdus merula

```{r, error=TRUE}
plot_individual_species_national(spec_name = "Turdus merula",
                                 spec_var_name = "Turdus.merula",
                                 model = mva_m1,
                                 at_list = list(unit_site="Mediterranean Maquis_Yagur"))
```

### Spilopelia senegalensis

```{r, error=TRUE}
plot_individual_species_national(spec_name = "Spilopelia senegalensis",
                                 spec_var_name = "Spilopelia.senegalensis",
                                 model = mva_m1,
                                 at_list = list(unit_site="Mediterranean Maquis_Givat Yeshayahu"))
```

### Parus major

```{r, error=TRUE}
plot_individual_species_national(spec_name = "Parus major",
                                 spec_var_name = "Parus.major",
                                 model = mva_m1,
                                 at_list = list(unit_site="Mediterranean Maquis_Givat Yeshayahu"))
```

### Passer domesticus

```{r, error=TRUE}
plot_individual_species_national(spec_name = "Passer domesticus",
                                 spec_var_name = "Passer.domesticus",
                                 model = mva_m1,
                                 disregard_p_val = TRUE)
```


```{r export manyglm p-values}
write.csv(coefp[,.(SciName,HebName,year_ct.p, unit_site.p)], file = file.path(output_path,"species_composition_p_values.csv"), row.names = F, fileEncoding = "MS-HEBR")
```

### Session information

```{r}
session_info()
```

