---
title: "Birds_Inland_Sands_analysis_by_plot"
author: "Ron Chen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# pcks <- list('devtools','rlang','ggplot2','data.table', 'dplyr','mvabund','kableExtra','lme4','MASS','vegan',
#              'jtools', 'interactions','car','ecoCopula','emmeans','ggrepel','ggvenn','Cairo')
# sapply(pcks, require, character = TRUE)
library (devtools)
library (rlang)
library (ggplot2)
library (data.table)
library (dplyr)
library (mvabund)
library (kableExtra)
library (lme4)
library (MASS)
library (vegan)
library (jtools)
library (interactions)
library (car)
library (ecoCopula)
library (emmeans)
library (ggrepel)
library (ggvenn)
library (Cairo)

source("plot_model_effect.R")
source("prepare_bird_data_with_Z_filter.R")
source("plot_alpha_diversity.R")
source("plot_coefs_with_traits.R")
source("plot_coefs_with_traits_for_publication.R")
source("multiplot.R")
source("plot_model_interaction.R")
# source_url("https://raw.githubusercontent.com/ronchen1/ron_functions/d2d0bab4dc65f29e53d316ae4098d0469711905b/R/multiplot.R")

# define font name
SoN_fontname <- "Almoni ML v5 AAA"

# Define the output folder path
output_path <- file.path("..","output", "inland_sands")

# Check if the output folder exists, and create it if it doesn't
if (!dir.exists(output_path)) {
  dir.create(output_path, recursive = TRUE)
  message("Folder '", output_path, "' has been created.")
} else {
  message("Folder '", output_path, "' already exists.")
}
```

```{r load using function}
out <- prepare_bird_data()
var_names <- names(out)
for (i in 1:length(var_names)) {
  assign(var_names[i], out[[i]])
}
rm(out)
P <- P_byplot[grepl("Inland",unit),][order(year,site)][,site:=factor(site)][,agriculture:=factor(agriculture)][,dunes:=factor(dunes)]
P_no_rare <- P_byplot_no_rare[grepl("Inland",unit),][order(year,site)][,site:=factor(site)][,agriculture:=factor(agriculture)][,dunes:=factor(dunes)]
```

## Inland Sands Bird analysis

Monitoring started in 2017 and thus there are three campaigns until today: 2017, 2019 and 2021. 
There are three sites: Beer Milka, Secher and Shunra East. 
Factors are:
 - proximity to agriculture (Far vs. Near): only in Beer Milka. Other two sites are only Far.
 - dune status (Shifting vs. Semi-Shifting): in all three sites.
There are a total of 12 plots, distributed as follows:

`r kable(plots_campaigns[grepl("Inland",unit),.(plot_num = uniqueN(point_name)),keyby = .(site,agriculture,dunes)])`


**Raw data**
Total abundance: `r Araw[grepl("Inland",unit),sum(total_count)]`
Number of observations: `r Araw[grepl("Inland",unit),.N]`
Total richness: `r Araw[grepl("Inland",unit),uniqueN(SciName)]`


**Filtered data**
Total abundance: `r A[grepl("Inland",unit),sum(count_under_250)]`
Number of observations: `r A[grepl("Inland",unit) & count_under_250>0,.N]`
Total richness: `r A[grepl("Inland",unit) & count_under_250>0,uniqueN(SciName)]`


#### ATTENTION: after filtration of rare species, there remain only `r A_no_rare[grepl("Inland",unit) & count_under_250>0, uniqueN(SciName)]` species. There are only `r A[grepl("Inland",unit) & count_under_250>0, sum(count_under_250),keyby = SciName][,sum(V1>10)]` species in this unit with abundance > 10.


### Model gma, abundance and richness

Richness done *with* rare species.
Abundance and mean abundance done *without* rare species.
Full models include cosine and sine of the time difference from June 21st, as well as time difference from sunrise (in radians).

#### richness

Explore data.

```{r richness data explore}
# include rare species in analysis
P.anal <- copy(P) # set a fixed variable name for analysis, if want to switch between data WITH rare species and data WITHOUT rare species then only change once here
print("RICHNESS WITH RARE SPECIES")
plot_alpha_diversity(P.anal, x_val = "agriculture", y_val = "richness", ylab_val = "richness", xlab_val = "agriculture")
plot_alpha_diversity(P.anal, x_val = "dunes", y_val = "richness", ylab_val = "richness", xlab_val = "dunes")
dotchart(P.anal$richness, groups = P.anal$agriculture, pch = as.numeric(P.anal$dunes), ylab = "agriculture", xlab = "richness")
legend(legend = levels(P.anal$dunes), x="topr", pch = as.numeric(unique(P.anal$dunes)))
dotchart(P.anal$richness, groups = P.anal$dunes, pch = as.numeric(P.anal$agriculture), ylab = "dunes", xlab = "richness")
legend(legend = levels(P.anal$agriculture), x=-1.6,y=10, pch = as.numeric(unique(P.anal$agriculture)))
IVs <- c("richness", "year_ct","site","agriculture", "dunes", "td_sc", "cos_td_rad", "sin_td_rad", "h_from_sunrise", "cos_hsun", "sin_hsun", "monitors_name",
                        "wind", "precipitation", "temperature", "clouds")
kable(summary(P.anal[,..IVs]))%>% kable_styling(font_size = 11)
```

observations missing for the 4 weather variables - exclude.
Include sampling time of day.


```{r richness data explore 2}
pairs(P.anal[,lapply(X = .SD,FUN = as.numeric),.SDcols=IVs[1:12]])
kable(cor(P.anal[,lapply(X = .SD,FUN = as.numeric),.SDcols=IVs[1:12]], use = "pairwise.complete.obs")) %>% kable_styling(font_size = 11)
```

Fit Poisson glm, check for existence of overdispersion

```{r model specification richness, warning=FALSE}
mdl_r.poiss.int <- glm(data = P.anal, formula = richness ~ agriculture * year_ct + dunes * year_ct + cos_td_rad + sin_td_rad + cos_hsun + sin_hsun + site, family = poisson)
od <- mdl_r.poiss.int$deviance/mdl_r.poiss.int$df.residual
print("Estimating overdispersion parameter phi: (Res. Dev.)/(n-p) where n=number of observations; p=number of parameters in the model.")
print(paste0('od = ',od))
```

Overdispersion parameter is < 1. Poisson more appropriate. Compare Poisson and negative binomial.

```{r model specification richness 2}
m0.po <- glm(data = P.anal, formula = richness ~ agriculture * year_ct + dunes * year_ct + cos_td_rad + sin_td_rad + cos_hsun + sin_hsun + site, family = poisson)
m0.nb <- glm.nb(data = P.anal, formula = richness ~ agriculture * year_ct + dunes * year_ct + cos_td_rad + sin_td_rad + cos_hsun + sin_hsun + site)
AIC(m0.po,m0.nb)
print("poisson")
plot(m0.po,which=1:3)
print("neg bin")
plot(m0.nb,which=1:3)
```

negative binomial did not converge because there is underdispersion and poisson is better.

```{r model specification richness 3}
m0 <- m0.po
me0 <- glmer(data = P.anal, formula = richness ~ agriculture * year_ct + dunes * year_ct + cos_td_rad + sin_td_rad + cos_hsun + sin_hsun + (1|site), family = poisson)
```

mixed model is singular. Go with glm.

```{r model specification richness 4}
summary(m0)
```

perform stepwise model selection of poisson model.

```{r model selection richness}
m1 <- step(m0)
```

Keep removing terms because $\Delta AIC<2$.
drop year:dunes

```{r model selection richness 2}
m1 <- glm(data = P.anal, formula = richness ~ agriculture + year_ct + dunes + sin_hsun, family = poisson)
drop1(m1)
```

drop dunes.

```{r}
m1 <- glm(data = P.anal, formula = richness ~ agriculture + year_ct + sin_hsun, family = poisson)
drop1(m1)
```

drop year.

```{r}
m1 <- glm(data = P.anal, formula = richness ~ agriculture + sin_hsun, family = poisson)
drop1(m1)
```

Final model includes agriculture and sampling time of day.
Final model:

```{r model selection richness 3}
summary(m1)
plot(m1, which=1:3)
```


```{r model interpretation richness}
summ(m1, exp=T, digits = 3)

plot_model_effect(P.anal = P.anal, m = m1, eff2plot = "agriculture", plot_points=TRUE, plot_residuals=FALSE, export_plot=TRUE, ylabel = NULL, fontname = SoN_fontname, fontsize=22, pdf_width=160, outpath = output_path)

effplot_agr <- effect_plot(model = m1, data=P.anal, pred = agriculture, interval = T, plot.points = T, jitter = c(0.1,0), point.size = 3, cat.pred.point.size = 4, partial.residuals = F, colors = "Qual1", main.title = "Effect of proximity to agriculture (points are observations)", line.colors = "black", point.alpha = 0.25)
effplot_agr$layers[[2]]$geom_params$width <- 0.4
effplot_agr
iv_agr_abs <- diff(effplot_agr$data$richness)
iv_agr_rel <- iv_agr_abs / as.data.table(effplot_agr$data)[agriculture=="Far",richness] * 100

m_coef <- coef(m1)
X <- unique(data.table(samp_time = P.anal$h_from_sunrise,
                       contrib = exp(m_coef["sin_hsun"]*P.anal$sin_hsun))[order(samp_time)])
plot(X$samp_time, X$contrib,xlab = "Hours from sunrise", ylab = "Effect on richness (multiplier)", type = "b")

# effect_plot(model = m1, data=P.anal, pred = sin_hsun, interval = T, plot.points = F, jitter = c(0.1,0), point.size = 3, cat.pred.point.size = 4, partial.residuals = T, colors = "Qual1", main.title = "Effect of sine of hours from sunrise (points are partial residuals)", line.colors = "black", point.alpha = 0.25)
```


**There is a statistically significant effect for proximity to agriculture.**

Near plots have on average `r iv_agr_abs` more species than far plots, which is `r iv_agr_rel` percent higher.

**No statistically significant effect for the other factors tested.**


#### geometric mean of abundance

#### ATTENTION: this is based on `r A_no_rare[grepl("Inland",unit) & count_under_250>0, uniqueN(SciName)]` species. There are only `r A[grepl("Inland",unit) & count_under_250>0, sum(count_under_250),keyby = SciName][,sum(V1>10)]` species in this unit with abundance > 10.####

Explore data.
Include time of day.

```{r gma data explore}
# exclude rare species in analysis
P.anal <- copy(P_no_rare)

print("GEOMETRIC MEAN ABUNDANCE WITHOUT RARE SPECIES")
plot_alpha_diversity(P.anal, x_val = "agriculture", y_val = "gma", ylab_val = "gma", xlab_val = "agriculture")
plot_alpha_diversity(P.anal, x_val = "dunes", y_val = "gma", ylab_val = "gma", xlab_val = "dunes")
dotchart(P.anal$gma, groups = P.anal$agriculture, pch = as.numeric(P.anal$dunes), ylab = "agriculture", xlab = "gma")
legend(legend = levels(P.anal$dunes), x="topr", pch = as.numeric(unique(P.anal$dunes)))
dotchart(P.anal$gma, groups = P.anal$dunes, pch = as.numeric(P.anal$agriculture), ylab = "dunes", xlab = "gma")
legend(legend = levels(P.anal$agriculture), x="topr", pch = as.numeric(unique(P.anal$agriculture)))
IVs <- c("gma", "year_ct","site","agriculture", "td_sc", "cos_td_rad", "sin_td_rad", "h_from_sunrise", "cos_hsun", "sin_hsun", "monitors_name",
                        "wind", "precipitation", "temperature", "clouds")
kable(summary(P.anal[,..IVs]))%>% kable_styling(font_size = 11)
```

extreme outlier where gma=19. Examine:

```{r}
print(P.anal[gma>15,])
```

Exclude this plot.
Fit glm, compare gamma, gaussian (poisson inappropriate because response is not discrete)

```{r model specification gma, warning=FALSE}
P.anal <- P.anal[gma<=18]
mdl_g.gamma.int <- glm(data = P.anal, formula = gma ~ agriculture*year_ct + dunes*year_ct + cos_td_rad + sin_td_rad + sin_hsun + cos_hsun + site, family = Gamma)
mdl_g.gauss.int <- glm(data = P.anal, formula = gma ~ agriculture*year_ct + dunes*year_ct + cos_td_rad + sin_td_rad + sin_hsun + cos_hsun + site, family = gaussian)
plot(mdl_g.gamma.int,which = 1:3,sub.caption = "gamma")
plot(mdl_g.gauss.int,which = 1:3,sub.caption = "gaussian")
```

Gamma seems better than gaussian.
Fit fixed and mixed models.

```{r model specification gma 2}
m0 <- glm(data = P.anal, formula = gma ~ agriculture*year_ct + dunes*year_ct + cos_td_rad + sin_td_rad + sin_hsun + cos_hsun + site, family = Gamma)
me0 <- glmer(data = P.anal, formula =  gma ~ agriculture*year_ct + dunes*year_ct + cos_td_rad + sin_td_rad + sin_hsun + cos_hsun + (1|site), family = Gamma)
```

Mixed model did not converge, use glm:

```{r model specification gma 3}
summary(m0)
```

perform stepwise model selection of Gamma model.

```{r model selection gma}
m1 <- step(m0)
```

keep removing terms, because $\Delta AIC<2$.

```{r}
drop1(m1)
```

drop agriculture:year

```{r}
m1 <- glm(data = P.anal, formula = gma ~ agriculture + year_ct + site + cos_td_rad, family = Gamma)
drop1(m1)
```

drop cosine.

```{r}
m1 <- glm(data = P.anal, formula = gma ~ agriculture + year_ct + site, family = Gamma)
drop1(m1)
```

drop year.

```{r}
m1 <- glm(data = P.anal, formula = gma ~ agriculture + site, family = Gamma)
drop1(m1)
```

keep removing terms, because $\Delta AIC<2$.
drop agriculture.

```{r}
m1 <- glm(data = P.anal, formula = gma ~ site, family = Gamma)
drop1(m1)
```

In the final model only site remains.

**No significant temporal or spatial trends detected in GMA.**

#### abundance

#### ATTENTION: this is based on `r A_no_rare[grepl("Inland",unit) & count_under_250>0, uniqueN(SciName)]` species. There are only `r A[grepl("Inland",unit) & count_under_250>0, sum(count_under_250),keyby = SciName][,sum(V1>10)]` species in this unit with abundance > 10.####

Explore data.

```{r abundance data explore}
# exclude rare species in analysis
P.anal <- copy(P_no_rare)

print("ABUNDANCE WITHOUT RARE SPECIES")
plot_alpha_diversity(P.anal, x_val = "agriculture", y_val = "abundance", ylab_val = "abundance", xlab_val = "agriculture")
plot_alpha_diversity(P.anal, x_val = "dunes", y_val = "abundance", ylab_val = "abundance", xlab_val = "dunes")
dotchart(P.anal$abundance, groups = P.anal$agriculture, pch = as.numeric(P.anal$dunes), ylab = "agriculture", xlab = "abundance")
legend(legend = levels(P.anal$dunes), x="topr", pch = as.numeric(unique(P.anal$dunes)))
dotchart(P.anal$abundance, groups = P.anal$dunes, pch = as.numeric(P.anal$agriculture), ylab = "dunes", xlab = "abundance")
legend(legend = levels(P.anal$agriculture), x=-1.6,y=10, pch = as.numeric(unique(P.anal$agriculture)))
IVs <- c("abundance", "year_ct","site","agriculture", "td_sc", "cos_td_rad", "sin_td_rad", "h_from_sunrise", "cos_hsun", "sin_hsun", "monitors_name",
                        "wind", "precipitation", "temperature", "clouds")
kable(summary(P.anal[,..IVs]))%>% kable_styling(font_size = 11)
```


```{r model specification abundance, warning=FALSE, eval=FALSE}
mdl_a.po <- glm(data = P.anal, formula = abundance ~ agriculture*year_ct + dunes*year_ct + cos_td_rad + sin_td_rad + sin_hsun + cos_hsun + site, family = poisson)
od <- mdl_a.po$deviance/mdl_a.po$df.residual
print("Estimating overdispersion parameter phi: (Res. Dev.)/(n-p) where n=number of observations; p=number of parameters in the model.")
print(paste0('od = ',od))
plot(mdl_a.po,which = 1:2,sub.caption = "poisson")
```

PHI>1, hence choose negative binomial.
Fit fixed and mixed models.
Choose mixed model if possible, otherwise choose a model with fixed-effects only.

```{r model specification abundance 2}
m0 <- glm.nb(data = P.anal, formula = abundance ~ agriculture*year_ct + dunes*year_ct + cos_td_rad + sin_td_rad + sin_hsun + cos_hsun + site)
me0 <- glmer.nb(data = P.anal, formula =  abundance ~ agriculture*year_ct + dunes*year_ct + cos_td_rad + sin_td_rad + sin_hsun + cos_hsun + (1|site))
```

mixed model is singular. go with fixed effects glm.

```{r model specification abundance 3}
summary(m0)
plot(m0, which=1:3)
```

Perform stepwise model selection of mixed model.

```{r model selection abundance}
m1 <- step(m0)
```

dunes * year, agriculture and sampling time of day remain.
The final model:

```{r model selection abundance 3}
summary(m1)
plot(m1, which=1:3)
```

Interpretation of abundance model:

```{r model interpretation abundance, warning=FALSE}
summ(m1,digits=3, exp=T)

plot_model_interaction(P.anal = P.anal, m = m1, eff2plot = "year_ct", modvar2plot = "dunes", plot_points=FALSE, plot_residuals=FALSE, export_plot=TRUE, ylabel = NULL, fontname = SoN_fontname, fontsize=22, pdf_width=160, outpath = output_path)

intplot_year_dunes <- interact_plot(model = m1, data=P.anal, pred = year_ct, modx = dunes, partial.residuals = F, jitter = c(0.2), point.size = 3, rug = F, point.shape = F, interval = T, main.title = "Interaction of time with dune status")
intplot_year_dunes

plot_model_effect(P.anal = P.anal, m = m1, eff2plot = "agriculture", export_plot = TRUE, plot_points = TRUE, fontname = SoN_fontname, outpath = output_path)

effplot_agr <- effect_plot(model = m1, data=P.anal, pred = agriculture, interval = T, plot.points = T, jitter = c(0.1,0), point.size = 3, cat.pred.point.size = 4, partial.residuals = F, colors = "Qual1", main.title = "Effect of proximity to agriculture (points are observations)", line.colors = "black", point.alpha = 0.25)
effplot_agr$layers[[2]]$geom_params$width <- 0.4
effplot_agr
iv_agr_abs <- diff(effplot_agr$data$abundance)
iv_agr_rel <- iv_agr_abs / as.data.table(effplot_agr$data)[agriculture=="Far",abundance] * 100

m_coef <- coef(m1)
X <- unique(data.table(samp_time = P.anal$h_from_sunrise,
                       contrib = exp(m_coef["cos_hsun"]*P.anal$cos_hsun))[order(samp_time)])
plot(X$samp_time, X$contrib,xlab = "Hours from sunrise", ylab = "Effect on abundance (multiplier)", type = "b")
```


Test for difference of temporal trend from zero for each dune status.

```{r model interpretation abundance 2}
emm_year_ct <- emtrends(m1, specs = "dunes", var = "year_ct", type = "response")
test_results_year_ct <- test(emm_year_ct, null = 0, adjust = "fdr")
print(test_results_year_ct)

emm_year_ct_dt <- as.data.table(emm_year_ct)
print(emm_year_ct_dt)
iv_shifting_year_rel <- emm_year_ct_dt[dunes=="Shifting",exp(year_ct.trend)]
print(iv_shifting_year_rel)
iv_semi_shifting_year_rel <- emm_year_ct_dt[grepl("Semi",dunes),exp(year_ct.trend)]
print(iv_semi_shifting_year_rel)
```

The trend of shifting sands is significantly different from zero, semi-shifting is not.

Test for pairwise differences in average abundance among levels of dunes, apply FDR correction for multiple comparisons.

```{r model interpretation abundance 3}
emm_dunes <- emmeans(object = m1, ~dunes*year_ct)
test_results_dunes <- test(pairs(emm_dunes, by="year_ct"), by=NULL, adjust="fdr")
print(test_results_dunes)

emm_dunes_dt <- as.data.table(emm_dunes)
iv_shifting_abs <- emm_dunes_dt[dunes=="Shifting",exp(emmean)]
iv_semi_shifting_abs <- emm_dunes_dt[grepl("Semi",dunes),exp(emmean)]
```

**There is a significant difference in average abundance among the two levels of distance from agriculture -  near plots have on average `r iv_agr_abs` more individuals than far plots, which is `r iv_agr_rel`%.**

**No effect of proximity to agriculture on the temporal trend of abundance.**

**No significant difference in average abundance among the two dune statuses -  shifting (`r iv_shifting_abs`) and semi-shifting (`r iv_semi_shifting_abs`).**

**The temporal trend of shifting sands is significantly different from zero (P=0.001), semi-shifting is not. The rate is as follows: shifting sands decrease `r (1-iv_shifting_year_rel)*100`% per year.**


#### community analysis using package MVabund

Not performed, because after filtration of rare species we are left with very few species.

#### ATTENTION: this is based on `r A_no_rare[grepl("Inland",unit) & count_under_250>0, uniqueN(SciName)]` species. There are only `r A[grepl("Inland",unit) & count_under_250>0, sum(count_under_250),keyby = SciName][,sum(V1>10)]` species in this unit with abundance > 10.####