---
title: "Birds_Maquis_analysis_by_plot_no_subunit_interactions"
author: "Ron Chen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
#knitr::opts_chunk$set(dev = "cairo_pdf", dev.args=list(cairo_pdf = list(family='Arial')))

# pcks <- list('devtools','rlang','ggplot2','data.table', 'dplyr','mvabund','kableExtra','lme4','MASS','vegan',
#              'jtools', 'interactions','car','ecoCopula','ggrepel','emmeans')
# sapply(pcks, require, character = TRUE)
library (devtools)
library (rlang)
library (ggplot2)
library (data.table)
library (dplyr)
library (mvabund)
library (kableExtra)
library (lme4)
library (MASS)
library (vegan)
library (jtools)
library (interactions)
library (car)
library (ecoCopula)
library (ggrepel)
library (emmeans)

source("plot_individual_species_maquis.R")
source("plot_model_effect.R")
source("prepare_bird_data_with_Z_filter.R")
source("plot_alpha_diversity.R")
source("plot_coefs_with_traits.R")
source("plot_coefs_with_traits_for_publication.R")
source("plot_two_coefs_with_traits.R")
source("multiplot.R")
# source_url("https://raw.githubusercontent.com/ronchen1/ron_functions/d2d0bab4dc65f29e53d316ae4098d0469711905b/R/multiplot.R")

# define font name
SoN_fontname <- "Almoni ML v5 AAA"

# Define the output folder path
output_path <- file.path("..","output", "maquis")

# Check if the output folder exists, and create it if it doesn't
if (!dir.exists(output_path)) {
  dir.create(output_path, recursive = TRUE)
  message("Folder '", output_path, "' has been created.")
} else {
  message("Folder '", output_path, "' already exists.")
}
```

```{r load using function}
out <- prepare_bird_data()
var_names <- names(out)
for (i in 1:length(var_names)) {
  assign(var_names[i], out[[i]])
}
rm(out)
P <- P_byplot[grepl("Maquis",unit),][order(year,site)][,site:=factor(site)][,subunit:=factor(subunit)]
P_no_rare <- P_byplot_no_rare[grepl("Maquis",unit),][order(year,site)][,subunit:=factor(subunit)]
```

## Mediterranean Maquis

Unit is divided into 3 subunits: Judea, Carmel and Galilee. Factors are proximity to settlements and time. Sampling started in spring 2012 (pilot year), but during T0 sampling was performed only in winter of 2014, and the next spring sampling was done in T1 (2015). Therefore 2012 will not be considered as pilot here but as T0. Total 5 campaigns, 3subunits per campaign, 5 sites per subunit, with 6 plots per site (total of 450 plot-campaign combinations).

**Raw data** Total abundance: `r Araw[grepl("aquis",unit),sum(total_count)]` Number of observations: `r Araw[grepl("aquis",unit),.N]` Total richness: `r Araw[grepl("aquis",unit),uniqueN(SciName)]`

**Filtered data** Total abundance: `r A[grepl("aquis",unit),sum(count_under_250)]` Number of observations: `r A[grepl("aquis",unit) & count_under_250>0,.N]` Total richness: `r A[grepl("aquis",unit) & count_under_250>0,uniqueN(SciName)]`

### Model gma, abundance and richness

Richness done *with* rare species. Abundance and mean abundance done *without* rare species. Full models include cosine and sine of the time difference from June 21st (in radians).

#### richness

Explore data and plot mean-variance plot. There is a strong relationship, indicating that employing GLMs is the proper way to analyze, rather than OLS (assumption of homogeneity is violated).

```{r richness data explore}
# include rare species in analysis
P.anal <- copy(P) # set a fixed variable name for analysis, if want to switch between data WITH rare species and data WITHOUT rare species then only change once here
print("RICHNESS WITH RARE SPECIES")
plot_alpha_diversity(P.anal, x_val = "settlements", y_val = "richness", ylab_val = "richness", xlab_val = "settlements")
plot_alpha_diversity(P.anal, x_val = "subunit", y_val = "richness", ylab_val = "richness", xlab_val = "subunit")

dotchart(P.anal$richness, groups = P.anal$settlements, pch = as.numeric(P.anal$settlements), ylab = "settlements", xlab = "richness")
dotchart(P.anal$richness, groups = P.anal$subunit, pch = as.numeric(P.anal$subunit), ylab = "subunit", xlab = "richness")
IVs <- c("richness", "year_ct","site","settlements","subunit", "td_sc", "cos_td_rad", "sin_td_rad", "h_from_sunrise", "cos_hsun", "sin_hsun", "monitors_name",
         "wind", "precipitation", "temperature", "clouds")
kable(summary(P.anal[,..IVs]))%>% kable_styling(font_size = 11)
```

no observation for all 4 weather variables. many NAs for sampling time of day variables.exclude from model. An extreme observation of richness\>20 in Judean highlands near settlements.

```{r richness data explore 3}
pairs(P.anal[,lapply(X = .SD,FUN = as.numeric),.SDcols=IVs[1:11]])
kable(cor(P.anal[,lapply(X = .SD,FUN = as.numeric),.SDcols=IVs[1:11]], use = "pairwise.complete.obs")) %>% kable_styling(font_size = 11)
```

Fit Poisson glm, check for existence of overdispersion

```{r model specification richness, warning=FALSE}
mdl_r.poiss.int <- glm(data = P.anal, formula = richness ~ settlements * year_ct + subunit +  cos_td_rad + sin_td_rad , family = poisson)
od <- mdl_r.poiss.int$deviance/mdl_r.poiss.int$df.residual
print("Estimating overdispersion parameter phi: (Res. Dev.)/(n-p) where n=number of observations; p=number of parameters in the model.")
print(paste0('od = ',od))
```

Overdispersion parameter is \< 1. Choose Poisson.

```{r model specification richness 2}
m0 <- mdl_r.poiss.int
me0 <- glmer(data = P.anal, formula =  richness ~ settlements * year_ct + subunit + cos_td_rad + sin_td_rad + (1|site), family = poisson)
```

```{r}
vif(me0)
```

go with mixed model, attempt model selection yet.

```{r model specification richness 3}
summary(me0)
```

perform stepwise model selection of poisson mixed model.

```{r model selection richness}
drop1(me0)
```

remove settlements X year.

```{r model selection richness 2}
me1 <- glmer(data = P.anal, formula =  richness ~ settlements + year_ct + subunit + cos_td_rad + sin_td_rad + (1|site), family = poisson)
drop1(me1)
```

drop year

```{r model selection richness 3}
me1 <- glmer(data = P.anal, formula =  richness ~ settlements + subunit + cos_td_rad + sin_td_rad + (1|site), family = poisson)
drop1(me1)
```

drop sine.

```{r model selection richness 4}
me1 <- glmer(data = P.anal, formula =  richness ~ settlements + subunit + cos_td_rad + (1|site), family = poisson)
drop1(me1)
```

Settlements, subunit and cosin time of year remain. Final model:

```{r model selection richness 6}
summary(me1)
ranef(me1)
dotplot(ranef(me1, condVar=TRUE))
plot(me1)
qqmath(me1)
# scale location plot
plot(me1, sqrt(abs(resid(.))) ~ fitted(.),
     type = c("p", "smooth"),
     main="scale-location",
     par.settings = list(plot.line =
                           list(alpha=1, col = "red",
                                lty = 1, lwd = 2)))
```

attempt to add year, see if significant:

```{r model selection richness 7}
summary(glmer(data = P.anal, formula =  richness ~ year_ct + settlements + subunit + cos_td_rad + (1|site), family = poisson))
```

year not significant, rightfully dropped. center time of year variable, highly correlated with intercept.

```{r model selection richness 8}
P.anal[,`:=`(cos_td_rad_c=cos_td_rad-mean(cos_td_rad),
             sin_td_rad_c=sin_td_rad-mean(sin_td_rad))]
me1 <- glmer(data = P.anal, formula =  richness ~ settlements + subunit + cos_td_rad_c + (1|site), family = poisson)
summary(me1)
```

```{r model interpretation richness}
P.anal <- P.anal
summ(me1, exp=T, digits = 3)
# effplot1 <- effect_plot(model = me1, data=P.anal, pred = settlements, interval = T, plot.points = T, jitter = c(0.1,0), point.size = 3, cat.pred.point.size = 4, partial.residuals = T, colors = "Qual1", main.title = "effect of proximity to settlements (points are partial residuals)", line.colors = "black", point.alpha = 0.25)
# effplot1$layers[[2]]$geom_params$width <- 0.4


effplot1 <- plot_model_effect(P.anal = P.anal, m = me1, eff2plot = "settlements", export_plot = TRUE, plot_points = TRUE, fontname = SoN_fontname, outpath = output_path)
effplot1

#Near is 46.84% higher
iv1_abs <- diff(effplot1$data$richness)
print(iv1_abs)
iv1_rel <- iv1_abs / as.data.table(effplot1$data)[settlements=="Far",richness] * 100
print(iv1_rel)

effplot1$data$richness
9.846757 / 6.705529 - 1

effplot2 <- effect_plot(model = me1, data=P.anal, pred = subunit, interval = T, plot.points = T, partial.residuals = T, jitter = c(0.1,0), point.size = 3, cat.pred.point.size = 4, colors = "Qual1", main.title = "effect of subunit (points are partial residuals)", line.colors = "black", point.alpha = 0.25)
effplot2$layers[[2]]$geom_params$width <- 0.4
effplot2

#emmeans
EMM <- emmeans(object = me1, ~subunit)
print(EMM)

# Carmel vs Galilee
(exp(2.18)/exp(1.89) - 1) * 100
# Carmel vs Judea 
(exp(2.18)/exp(2.1) - 1) * 100
#Judea vs Galilee
(exp(2.1)/exp(1.89) - 1) * 100

test_results_subunit <- test(pairs(EMM, by=NULL, adjust="fdr"))
print(test_results_subunit)

```

**statistically significant lower richness in Galilee and far from settlements. No significant change in richness over time.**


#### geometric mean of abundance

Explore data. Exclude time of day because of high number of NAs.

```{r gma data explore}
# exclude rare species in analysis
P.anal <- copy(P_no_rare)
P.anal <- P.anal[pilot==FALSE]

print("GEOMETRIC MEAN ABUNDANCE WITHOUT RARE SPECIES")
plot_alpha_diversity(P.anal, x_val = "settlements", y_val = "gma", ylab_val = "gma", xlab_val = "settlements")
plot_alpha_diversity(P.anal, x_val = "subunit", y_val = "gma", ylab_val = "gma", xlab_val = "subunit")

dotchart(P.anal$gma, groups = P.anal$settlements, pch = as.numeric(P.anal$settlements), ylab = "settlements", xlab = "gma")
dotchart(P.anal$gma, groups = P.anal$subunit, pch = as.numeric(P.anal$subunit), ylab = "subunit", xlab = "gma")
IVs <- c("gma", "year_ct","site","settlements","subunit", "td_sc", "cos_td_rad", "sin_td_rad")
kable(summary(P.anal[,..IVs]))%>% kable_styling(font_size = 11)
```

Fit glm, compare gamma, gaussian (poisson inappropriate because response is not discrete)

```{r model specification gma, warning=FALSE}
# mdl_g.gamma.int <- glm(data = P.anal, formula = gma ~ settlements * year_ct + subunit*year_ct + cos_td_rad + sin_td_rad, family = Gamma) # run this if there are no zeros 
# plot(mdl_g.gamma.int,which = 1:3,sub.caption = "gamma")
mdl_g.gauss.int <- glm(data = P.anal, formula = gma ~ settlements * year_ct + subunit + cos_td_rad + sin_td_rad, family = gaussian)
plot(mdl_g.gauss.int,which = 1:3,sub.caption = "gaussian")
```

Remove rows 211, 376, 408. Fit fixed and mixed models.

```{r model specification gma 2}
P.anal <- P.anal[!c(211, 376, 408)]
m0 <- lm(data = P.anal, formula = gma ~ settlements * year_ct + subunit + cos_td_rad + sin_td_rad)
me0 <- lmer(data = P.anal, formula =  gma ~ settlements * year_ct + subunit + cos_td_rad + sin_td_rad + (1|site))
```

Mixed model converged

```{r model specification gma 3}
summary(me0)
```

perform stepwise model selection of gaussian model.

```{r model selection gma}
drop1(me0)
```

drop subunit

```{r}
me1 <- lmer(data = P.anal, formula =  gma ~ settlements * year_ct + cos_td_rad + sin_td_rad + (1|site))
drop1(me1)
```

drop year X settlements

```{r}
me1 <- lmer(data = P.anal, formula =  gma ~ settlements + year_ct + cos_td_rad + sin_td_rad + (1|site))
drop1(me1)
```

drop year

```{r}
me1 <- lmer(data = P.anal, formula =  gma ~ settlements + cos_td_rad + sin_td_rad + (1|site))
drop1(me1)
```

This is the final model:

```{r model selection gma 2}
summary(me1)
ranef(me1)
dotplot(ranef(me1, condVar=TRUE))
plot(me1)
qqmath(me1)
# scale location plot
plot(me1, sqrt(abs(resid(.))) ~ fitted(.),
     type = c("p", "smooth"),
     main="scale-location",
     par.settings = list(plot.line =
                           list(alpha=1, col = "red",
                                lty = 1, lwd = 2)))
```

**Not a great fit. settlement is not significant.**

```{r model interpretation gma}
summ(me1, digits = 3)
plot(me1, which=1:3)


effplot1 <- plot_model_effect(P.anal = P.anal, m = me1, eff2plot = "settlements", export_plot = TRUE, plot_points = TRUE, fontname = SoN_fontname, outpath = output_path, y_cutoff = 5)
effplot1

perc_diff <- diff(as.data.table(effplot1$data)[,gma])/as.data.table(effplot1$data)[settlements=="Far",gma]*100

# EMM <- emmeans(object = me1, ~ settlements)
# print(EMM)
# test_results_settlements <- test(pairs(EMM, by=NULL, adjust="fdr"))
# print(test_results_settlements)

```

**No significant effect in GMA.**

#### abundance

Explore data

```{r abundance data explore}
# exclude rare species in analysis
P.anal <- copy(P_no_rare)

print("ABUNDANCE WITHOUT RARE SPECIES")
plot_alpha_diversity(P.anal, x_val = "settlements", y_val = "abundance", ylab_val = "abundance", xlab_val = "settlements")
plot_alpha_diversity(P.anal, x_val = "subunit", y_val = "abundance", ylab_val = "abundance", xlab_val = "subunit")

dotchart(P.anal$abundance, groups = P.anal$settlements, pch = as.numeric(P.anal$settlements), ylab = "settlements", xlab = "abundance")
dotchart(P.anal$abundance, groups = P.anal$subunit, pch = as.numeric(P.anal$subunit), ylab = "subunit", xlab = "abundance")
IVs <- c("abundance", "year_ct","site","settlements","subunit", "td_sc", "cos_td_rad", "sin_td_rad")
kable(summary(P.anal[,..IVs]))%>% kable_styling(font_size = 11)
```

Some outliers with total abundance \>100. Examine:

```{r abundance data explore 2}
P.anal[abundance>100,.(subunit,point_name,datetime,monitors_name,richness,gma,abundance)]
```

Exclude 1 plot with high abundance (\>150) to improve model fit.

```{r model specification abundance, warning=FALSE, eval=FALSE}
P.anal <- P.anal[abundance<150]
mdl_a.po <- glm(data = P.anal, formula = abundance ~ settlements * year_ct + subunit + cos_td_rad + sin_td_rad, family = poisson)
od <- mdl_a.po$deviance/mdl_a.po$df.residual
print("Estimating overdispersion parameter phi: (Res. Dev.)/(n-p) where n=number of observations; p=number of parameters in the model.")
print(paste0('od = ',od))
plot(mdl_a.po,which = 1:2,sub.caption = "poisson")
```

PHI\>1, hence choose negative binomial. Fit fixed and mixed models. Choose mixed model if possible, otherwise choose a model with fixed-effects only.

```{r model specification abundance 2}
m0 <- glm.nb(data = P.anal, formula = abundance ~ settlements * year_ct + subunit + cos_td_rad + sin_td_rad)
AIC(m0)
me0 <- glmer.nb(data = P.anal, formula =  abundance ~ settlements * year_ct + subunit + cos_td_rad + sin_td_rad + (1|site))
AIC(me0)
```

mixed model converged, slightly better AIC. Perform stepwise model selection of mixed model.

```{r model selection abundance}
drop1(me0)
```

drop settlements X year

```{r model selection abundance 2}
me1 <- glmer.nb(data = P.anal, formula =  abundance ~ settlements + subunit + year_ct + cos_td_rad + sin_td_rad + (1|site))
drop1(me1)
```

drop year

```{r}
me1 <- glmer.nb(data = P.anal, formula =  abundance ~ settlements + subunit + cos_td_rad + sin_td_rad + (1|site))
drop1(me1)
```

Subunit, settlement and time of year remain. The final model:

```{r model selection abundance 3}
summary(me1)
ranef(me1)
dotplot(ranef(me1, condVar=TRUE))
plot(me1)
qqmath(me1)
# scale location plot
plot(me1, sqrt(abs(resid(.))) ~ fitted(.),
     type = c("p", "smooth"),
     main="scale-location",
     par.settings = list(plot.line =
                           list(alpha=1, col = "red",
                                lty = 1, lwd = 2)))
```

Not a great fit, high residuals. center time of year variables, highly correlated with intercept.

```{r}
P.anal[,`:=`(cos_td_rad_c=cos_td_rad-mean(cos_td_rad),
             sin_td_rad_c=sin_td_rad-mean(sin_td_rad))]
me1 <- glmer.nb(data = P.anal, formula =  abundance ~ settlements + subunit + cos_td_rad_c + sin_td_rad_c + (1|site))
summary(me1)
```

Interpretation of abundance model:

```{r model interpretation abundance, warning=FALSE}
summ(me1,exp=T,digits=3)

effplot1 <- effect_plot(model = me1, data=P.anal, pred = settlements, interval = T, plot.points = T, jitter = c(0.1,0), point.size = 3, cat.pred.point.size = 4, partial.residuals = F, colors = "Qual1", main.title = "effect of settlement proximity (points are  observations)", line.colors = "black", point.alpha = 0.25) + ylim(0,150)

effplot1


plot_model_effect(P.anal = P.anal, m = me1, eff2plot = "settlements", export_plot = TRUE, plot_points = TRUE, fontname = SoN_fontname, outpath = output_path, y_cutoff = 80)

#Far vs Near - near is 72.59% higher than far

print(effplot1$data$abundance)
iv1_abs <- diff(effplot1$data$abundance)
iv1_rel <- iv1_abs / as.data.table(effplot1$data)[settlements=="Far",abundance] * 100
print(iv1_rel)


plot_model_effect(P.anal = P.anal, m = me1, eff2plot = "subunit", export_plot = TRUE, plot_points = TRUE, fontname = SoN_fontname, outpath = output_path, y_cutoff = 80)

#emmeans
EMM <- emmeans(object = me1, ~ subunit)
print(EMM)
EMMdt <- as.data.table(EMM)

print("Carmel vs Galilee")
(exp(EMMdt[subunit %like% 'Carmel',emmean])/exp(EMMdt[subunit %like% 'Galil',emmean]) - 1) * 100
print("Carmel vs Judea ")
(exp(EMMdt[subunit %like% 'Carmel',emmean])/exp(EMMdt[subunit %like% 'Judea',emmean]) - 1) * 100
print("Judea vs Galilee")
(exp(EMMdt[subunit %like% 'Judea',emmean])/exp(EMMdt[subunit %like% 'Galil',emmean]) - 1) * 100

test_results_subunit <- test(pairs(EMM, by=NULL, adjust="fdr"))
print(test_results_subunit)

```

**significantly lower abundance in Galilee subunit and far from settlements. No significant temporal trend.**

#### community analysis using package MVabund

```{r explore single observations}
abu_by_spp.maquis <- abu_by_spp[grepl("Maquis",unit),]
abu_by_spp_no_rare.maquis <- abu_by_spp_no_rare[grepl("Maquis",unit),]

spp <- abu_by_spp.maquis[,.SD[,(length(col_names)+1):ncol(abu_by_spp.maquis)]]
# filter out species with zero counts (were either not observed or are rare and were excluded)
spp <- spp[,.SD,.SDcols = colSums(spp)>0]
spp <- mvabund(spp)

spp_no_rare <- abu_by_spp_no_rare.maquis[,.SD[,(length(col_names)+1):ncol(abu_by_spp_no_rare.maquis)]]
# filter out species with zero counts (were either not observed or are rare and were excluded)
spp_no_rare <- spp_no_rare[,.SD,.SDcols = colSums(spp_no_rare)>0]
spp_no_rare <- mvabund(spp_no_rare)

env_data <- abu_by_spp_no_rare.maquis[pilot==FALSE,..col_names]
env_data[,site:=factor(site)][,subunit:=factor(subunit)]

try(expr = plot(spp ~ abu_by_spp.maquis$settlements,overall.main="raw abundances", transformation = "no"))
try(plot(spp ~ abu_by_spp.maquis$subunit,overall.main="raw abundances", transformation = "no"))

plot(sort(spp_no_rare[spp_no_rare>0]))
dotchart(sort(A[grepl("Maquis",unit) & count_under_250>=30,count_under_250]))
meanvar.plot(spp, xlab = "mean abundance of a given species across sites", ylab = "variance of the abundance of a given species across sites")
```

There are few observations with counts of \>60. Examine these:

```{r explore single observations 2}
A[grepl("Maquis",unit) & count_under_250>=60,.(point_name,datetime,SciName,monitors_name,count_under_250)]
```

Both observations are of P. domesticus, it is likely to be seen in large groups.

**Look for species that were completely absent in certain subunits or settlemnt proximities:**

```{r explore observations}
A_no_rare[unit %like% "Maquis",sum(count_under_250), keyby = .(SciName, subunit)][,is_zero:=V1==0][is_zero==TRUE] %>% kbl() %>%  kable_styling(font_size = 11)
A_no_rare[unit %like% "Maquis",sum(count_under_250), keyby = .(SciName, settlements)][,is_zero:=V1==0][is_zero==TRUE] %>% kbl() %>%  kable_styling(font_size = 11)
```

**Curruca curruca was not observed in Judean Highlands at all. Validate model specifically.**

start model specification:

```{r community analysis MVabund model specification}
mva_m0.po <- manyglm(formula = spp_no_rare ~ settlements*year_ct + subunit + cos_td_rad + sin_td_rad, family = "poisson", data = env_data)
mva_m0.nb <- manyglm(formula = spp_no_rare ~ settlements*year_ct + subunit + cos_td_rad + sin_td_rad, family = "negative.binomial", data = env_data)
aic_dt <- data.table(nb = mva_m0.nb$aic, po=mva_m0.po$aic)
colMeans(aic_dt)
print("POISSON")
plot(mva_m0.po, which=1:3)
print("NEGATIVE BINOMIAL")
plot(mva_m0.nb, which=1:3)
```

negative binomial model is better than poisson according to residuals and AIC comparison.

```{r community analysis MVabund model specification 2}
mva_m0 <- mva_m0.nb
mva_m1 <- manyglm(formula = spp_no_rare ~ settlements*year_ct + subunit + cos_td_rad + sin_td_rad + site, data = env_data)
aic_dt$nb1 <- mva_m1$aic
colMeans(aic_dt)
```

The addition of the explanatory variable 'site' is somewhat improving the AIC of the model. Prefer to exclude site, for simplification. stepwise selection of model:

```{r community analysis MVabund model selection}
drop1(mva_m0)
```

drop settlements X year.

```{r community analysis MVabund model selection 2}
mva_m1 <- manyglm(formula = spp_no_rare ~ settlements + subunit + year_ct + cos_td_rad + sin_td_rad, family = "negative.binomial",      data = env_data) 
drop1(mva_m1)
```

final model includes settlements, year, subunit, sampling time of year.

```{r community analysis MVabund model validation}
plot(mva_m1, which=1:3)
```

```{r community analysis MVabund model interpretation}
# summ_m1 <- summary(mva_m1)
# saveRDS(summ_m1, file = "Maquis_new_MVabund_summary_output_999iter.rds")
# anov_m1.uni <- anova(mva_m1, p.uni = "adjusted", nBoot = 99, show.time = "all")
# saveRDS(anov_m1.uni, file = "Maquis_new_MVabund_anova_output_99iter.rds")
summ_m1 <- readRDS(file = "../output/20240104_Maquis_new_MVabund_summary_output_999iter.rds")
anov_m1.uni <- readRDS("../output/20240104_Maquis_new_MVabund_anova_output_99iter.rds")
print(summ_m1)
print(anov_m1.uni)
```

**Factors settlements, year, subunit, time of year have a statistically significant effect on community composition.**

Check Curruca curruca model:
```{r}
d.curcur <- cbind(env_data,setnames(data.table(spp_no_rare[,"Curruca.curruca"]), old = 1, new = "Curruca.curruca"))
m.curcur <- glm.nb(formula = Curruca.curruca ~ settlements + subunit + year_ct + cos_td_rad + sin_td_rad, data = d.curcur)
summary(m.curcur)
```

subunit factor still significant, year variable not significant. replace p_value for Curruca curruca in manyglm results with p-value obtained from individual glm.

```{r}
anov_m1.uni$uni.p["year_ct","Curruca.curruca"] <- summary(m.curcur)$coefficients["year_ct","Pr(>|z|)"]
```

```{r community analysis MVabund model interpretation - plot coefficients}
coefp <- merge(data.table(t(coef(mva_m1)/log(2)),keep.rownames=TRUE), data.table(t(anov_m1.uni$uni.p), keep.rownames = TRUE), by = "rn") # coefficients are on log2 scale to simplify interpretation, yet keep presenetation on log scale rather than linear. The link function used for poisson is log -> above 0 is a positive trend and below 0 is negative trend

# add total species abundance
coefp <- merge(coefp,as.data.table(colSums(spp_no_rare),keep.rownames = TRUE), by.x = "rn", by.y = "V1")

# add hebrew name and traits
sci_heb <- unique(A[,.(SciName,HebName)])
sci_heb[,SciName:=make.names(SciName)]
coefp <- merge(coefp, sci_heb, by.x = "rn", by.y = "SciName")

# set column names based on the predictor variables
colnames(coefp)[grepl("rn",colnames(coefp))] <- "SciName"
colnames(coefp)[grepl("Intercept).x",colnames(coefp))] <- "intercept.coef"
colnames(coefp)[grepl("Intercept).y",colnames(coefp))] <- "intercept.p"
colnames(coefp)[colnames(coefp)=="settlementsNear"] <- "settlementsNear.coef"
colnames(coefp)[colnames(coefp)=="settlements"] <- "settlements.p"
colnames(coefp)[grepl("cos_td_rad.x",colnames(coefp))] <- "cos_td_rad.coef"
colnames(coefp)[colnames(coefp)=="cos_td_rad.y"] <- "cos_td_rad.p"
colnames(coefp)[grepl("sin_td_rad.x",colnames(coefp))] <- "sin_td_rad.coef"
colnames(coefp)[grepl("sin_td_rad.y",colnames(coefp))] <- "sin_td_rad.p"
colnames(coefp)[colnames(coefp)=="year_ct.x"] <- "year_ct.coef"
colnames(coefp)[colnames(coefp)=="year_ct.y"] <- "year_ct.p"
colnames(coefp)[colnames(coefp)=="V2"] <- "species_abundance"
colnames(coefp)[colnames(coefp)=="dunesshifting"] <- "dunesShifting.coef"
colnames(coefp)[colnames(coefp)=="dunes"] <- "dunes.p"
colnames(coefp)[colnames(coefp)=="site"] <- "site.p"
colnames(coefp)[colnames(coefp)=="subunitCarmel"] <- "subunitCarmel.coef"
colnames(coefp)[colnames(coefp)=="subunitGalilee"] <- "subunitGalilee.coef"
colnames(coefp)[colnames(coefp)=="subunit"] <- "subunit.p"


# plot_coefs_settle <- plot_coefs_with_traits(D = A_no_rare[grepl("Maquis",unit)],
#                                      cp = coefp[,.(SciName,settlementsNear.coef,settlements.p,species_abundance)],
#                                      plot_title = "Effect of proximity to settlements on species abundance",
#                                      plot_xlabel = expression(paste("coefficient of NEAR settlements, (log"[2]," scale)")),
#                                      mark_batha = FALSE)
# plot_coefs_settle

plot_coefs_settle <- plot_coefs_with_traits_for_publication(D = A_no_rare[grepl("Maquis",unit)],
                                                            cp = coefp[,.(SciName,settlementsNear.coef,settlements.p,species_abundance)],
                                                            plot_title = "Effect of proximity to settlement on species abundance",
                                                            plot_xlabel = expression(paste("coefficient of NEAR settlement, (log"[2]," scale)")),
                                                            mark_batha = TRUE,
                                                            show_legend = FALSE,
                                                            export_plot = TRUE,
                                                            fontname = SoN_fontname,
                                                            effect_str = "settlements",
                                                            outpath = output_path)
plot_coefs_settle


# compute mean of year coefficients (geometric mean of linear scale coefficients) to plot single year coefficient
#coefp[,mn_year.coef:=year_ct.coef+`subunitCarmel:year_ct`/3+`subunitGalilee:year_ct`/3]
# if any of the relevant terms - year_ct, subunit X year - is <0.1, set the p-value of the mean year coef to be 0.01 (arbitrary, lower than 0.1 which is the threshold)
#coefp[,mn_year.p:=apply(X = .SD, 1, FUN = \(x) 1-sum(x<0.1)*0.99), .SDcols=c("year_ct.p","subunit:year_ct")]

# plot_coefs_year <- plot_coefs_with_traits(D = A_no_rare[grepl("Maquis",unit)],
#                                      cp = coefp[,.(SciName,mn_year.coef,mn_year.p,species_abundance)],
#                                      plot_title = "Mean temporal effect on species abundance",
#                                      plot_xlabel = expression(paste("coefficient of year counter, (log"[2]," scale)")),
#                                      mark_batha = FALSE)


plot_coefs_year <- plot_coefs_with_traits_for_publication(D = A_no_rare[grepl("Maquis",unit)],
                                                          cp = coefp[,.(SciName,year_ct.coef,year_ct.p,species_abundance)],
                                                          plot_title = "Temporal effect on species abundance",
                                                          plot_xlabel = expression(paste("coefficient of year counter, (log"[2]," scale)")),
                                                          mark_batha = TRUE,
                                                          show_legend = FALSE,
                                                          export_plot = TRUE,
                                                          fontname = SoN_fontname,
                                                          effect_str = "year_ct",
                                                          outpath = output_path)

plot_coefs_year

# plot_coefs_subunit <- plot_coefs_with_traits_for_publication(D = A_no_rare[grepl("Maquis",unit)],
#                                      cp = coefp[,.(SciName,subunitCarmel.coef,subunit.p,species_abundance)],
#                                      plot_title = "Effect of sub-unit on species abundance",
#                                      plot_xlabel = expression(paste("coefficient of Carmel sub-unit, (log"[2]," scale)")),
#                                      mark_batha = TRUE,
#                                                 show_legend = FALSE,
#                                                 export_plot = TRUE,
#                                                 fontname = SoN_fontname,
#                                                 effect_str = "subunit",
#                                                 outpath = output_path)

plot_coefs_subunit <- plot_two_coefs_with_traits(D = A_no_rare[grepl("Maquis",unit)],
                                                 cp = coefp[,.(SciName,coef.x=subunitCarmel.coef,coef.y=subunitGalilee.coef,subunit.p,species_abundance)],
                                                 plot_title = "Effect of sub-unit on species abundance",
                                                 plot_xlabel = expression(paste("coefficient of Carmel sub-unit, (log"[2]," scale)")),
                                                 plot_ylabel = expression(paste("coefficient of Galilee sub-unit, (log"[2]," scale)")),
                                                 mark_batha = TRUE)

plot_coefs_subunit

# plot the coefficients of year considering interaction with subunits
# coef_cutoff <- 0.25 # do not display species with all three coefs below this
# X <- coefp[,.(Judea=year_ct.coef,
#               Carmel=year_ct.coef + `subunitCarmel:year_ct`,
#               Galil=year_ct.coef + `subunitGalilee:year_ct`,
#               species_abundance,
#               SciName,HebName)]
# Xcut <- X[abs(Judea)>coef_cutoff | abs(Carmel)>coef_cutoff | abs(Galil)>coef_cutoff]
# col_scale_min <- -0.7
# p_yearsubu <- ggplot(data = Xcut, aes(x = Galil, y = Carmel, color=Judea)) +
#   theme_light() +
#   geom_hline(yintercept = 0, color = "darkslategrey") +
#   geom_vline(xintercept = 0, color = "darkslategrey") +
#   geom_point(size=3.5) +
#   scale_color_gradient2(midpoint=0, low="darkred", mid="white",high="darkblue", space ="Lab",
#                         limits=c(col_scale_min,NA), oob=scales::squish, name=paste0("Judea (lower limit=",col_scale_min,")")) +
#   geom_text_repel(aes(label = paste0("(",species_abundance,") ",HebName)), color="black", size=3.5, force_pull = 0, max.overlaps = 50,
#                   max.iter = 1e5, box.padding = 0.7, min.segment.length = 0) +
#   ggtitle("Year coefficient for three Mediterranean Maquis subunits")
# 
# p_yearsubu

```

### Alectoris chukar - חוגלת סלעים

```{r community analysis MVabund interpretation alechu, error=TRUE}
plot_individual_species_maquis(spec_name = "Alectoris chukar", spec_var_name = "Alectoris.chukar")
```

### Acridotheres tristis -  מיינה מצויה

```{r community analysis MVabund interpretation acrtri, error=TRUE}
plot_individual_species_maquis(spec_name = "Acridotheres tristis", spec_var_name = "Acridotheres.tristis")
```

### Streptopelia turtur - תור מצוי

```{r community analysis MVabund interpretation strtur, error=TRUE}
plot_individual_species_maquis(spec_name = "Streptopelia turtur", spec_var_name = "Streptopelia.turtur")
```

### Columba livia - יונת סלעים (ותת-מין יונת הבית)

```{r community analysis MVabund interpretation colliv, error=TRUE}
plot_individual_species_maquis(spec_name = "Columba livia", spec_var_name = "Columba.livia")
```

### Passer domesticus - דרור הבית

```{r community analysis MVabund interpretation pasdom, error=TRUE}
plot_individual_species_maquis(spec_name = "Passer domesticus", spec_var_name = "Passer.domesticus")
```

### Spilopelia senegalensis - צוצלת

```{r community analysis MVabund interpretation spisen, error=TRUE}
plot_individual_species_maquis(spec_name = "Spilopelia senegalensis", spec_var_name = "Spilopelia.senegalensis")
```

###  Cecropis daurica - סנונית מערות

```{r community analysis MVabund interpretation cecdau, error=TRUE}
plot_individual_species_maquis(spec_name = "Cecropis daurica", spec_var_name = "Cecropis.daurica")
```

###  Psittacula krameri - דררה

```{r community analysis MVabund interpretation psikra, error=TRUE}
plot_individual_species_maquis(spec_name = "Psittacula krameri", spec_var_name = "Psittacula.krameri")
```

###  Prinia gracilis - פשוש

```{r community analysis MVabund interpretation prigra, error=TRUE}
plot_individual_species_maquis(spec_name = "Prinia gracilis", spec_var_name = "Prinia.gracilis")
```

###  Curruca curruca - סבכי טוחנים

```{r community analysis MVabund interpretation curcur, error=TRUE}
spec_name <- "Curruca curruca"
spec_var_name <- "Curruca.curruca"
at_list <- list(subunit = "Galilee")
y_cutoff <- NULL
plot_alpha_diversity(P = abu_by_spp.maquis, x_val = "settlements", y_val = spec_name, ylab_val = spec_name)
  d <- cbind(env_data,setnames(data.table(spp_no_rare[,spec_var_name]), old = 1, new = spec_var_name))
  d <- d[!(subunit %like% "Judea")]
  d[,subunit:=factor(subunit)]
  m <- glm.nb(formula = paste(spec_var_name,"settlements + subunit + year_ct + cos_td_rad + sin_td_rad", sep = "~"), data = d)
  print(coef(mva_m1)[,spec_var_name])
  print(coef(m))
  
  print(paste("RESULTS FOR",spec_name))
  print("P-values")
  pdt <- setnames(x = as.data.table(anov_m1.uni$uni.p, keep.rownames = T)[,.SD,.SDcols = c("rn",spec_var_name)], old = 1:2, new = c("term","p"))
  print(pdt)
  
  #Temporal trend----
  if (pdt[term=="year_ct",p<0.1]) {
    plot_time <- effect_plot(data = d, model = m, pred = year_ct, interval = T, plot.points = T, jitter = c(0.1,0), point.size = 3, cat.pred.point.size = 4,
                             partial.residuals = F, colors = "Qual1", main.title = spec_name, at = at_list)
    plot_time
    plot_model_effect(P.anal = d, m = m, eff2plot = "year_ct", export_plot = TRUE, plot_points = F, fontname = SoN_fontname,
                      outpath = output_path, x_breaks=seq(0,9), xticklabs=c("2012","","2014","","2016","","2018","","2020",""), at_list = at_list)
    time_dt <- as.data.table(plot_time$data)[,..spec_var_name]
    print(paste("change in",spec_name,"abundance over monitoring period:",(time_dt[nrow(time_dt)]-time_dt[1])/time_dt[1] * 100,"%"))
  }
  
  # spatial effect - settlements ----
  if (pdt[term=="settlements",p<0.1]) {
    plot_set <- effect_plot(data = d, model = m, pred = settlements, interval = T, plot.points = T, jitter = c(0.1,0), point.size = 3, cat.pred.point.size = 4,
                            partial.residuals = F, colors = "Qual1", main.title = spec_name, at = at_list)
    plot_set
    plot_model_effect(P.anal = d, m = m, eff2plot = "settlements", export_plot = TRUE, plot_points = T, fontname = SoN_fontname,
                      outpath = output_path, y_cutoff = y_cutoff, at_list = at_list)
    set_dt <- as.data.table(plot_set$data)[,..spec_var_name]
    print(paste(spec_name,"abundance in high proximity is",(max(set_dt)/min(set_dt) - 1) *100,"% higher than low proximity."))
  }
  
  # spatial effect - subunit----
  if (pdt[term=="subunit",p<0.1]) {
    plot_subunit <- effect_plot(data = d, model = m, pred = subunit, interval = T, plot.points = T, jitter = c(0.1,0), point.size = 3, cat.pred.point.size = 4,
                                partial.residuals = F, colors = "Qual1", main.title = spec_name, at = at_list)
    plot_subunit
    
    plot_model_effect(P.anal = d, m = m, eff2plot = "subunit", export_plot = TRUE, plot_points = T, fontname = SoN_fontname,
                      outpath = output_path, y_cutoff = y_cutoff, at_list = at_list)
    
    # post-hoc: test for pairwise differences in subunits if subunit term is significant
    EMM <- emmeans(object = m, ~ subunit)
    print(EMM)
    EMMdt <- as.data.table(EMM)
    
    print("Carmel vs Galilee (% difference)")
    print((exp(EMMdt[subunit %like% 'Carmel',emmean])/exp(EMMdt[subunit %like% 'Galil',emmean]) - 1) * 100)
    print("Carmel vs Judea (% difference)")
    print((exp(EMMdt[subunit %like% 'Carmel',emmean])/exp(EMMdt[subunit %like% 'Judea',emmean]) - 1) * 100)
    print("Galilee vs Judea (% difference)")
    print((exp(EMMdt[subunit %like% 'Galil',emmean])/exp(EMMdt[subunit %like% 'Judea',emmean]) - 1) * 100)
    
    test_results_subunit <- test(pairs(EMM, by=NULL, adjust="fdr"))
    print(test_results_subunit)
  }
  rm(spec_name)
  rm(spec_var_name)
  rm(d)
  rm(m)
  rm(pdt)
```

### Chloris chloris - ירקון

```{r community analysis MVabund interpretation chlchl, error=TRUE}
plot_individual_species_maquis(spec_name = "Chloris chloris", spec_var_name = "Chloris.chloris")
```

```{r export manyglm p-values}
write.csv(coefp[,.(SciName,HebName,year_ct.p, settlements.p, subunit.p)], file = "../output/maquis/species_composition_p_values_no_subunit_interaction.csv", row.names = F, fileEncoding = "MS-HEBR")
```