---
title: "Birds_Batha_analysis_by_plot"
author: "Ron Chen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
#knitr::opts_chunk$set(dev = "cairo_pdf", dev.args=list(cairo_pdf = list(family='Arial')))

pcks <- list('devtools','rlang','ggplot2','data.table', 'dplyr','mvabund','kableExtra','lme4','MASS','vegan',
             'jtools', 'interactions','car','ecoCopula','Cairo','extrafont', 'ggnewscale')
sapply(pcks, require, character = TRUE)
library (devtools)
library (rlang)
library (ggplot2)
library (data.table)
library (dplyr)
library (mvabund)
library (kableExtra)
library (lme4)
library (MASS)
library (vegan)
library (jtools)
library (interactions)
library (car)
library (ecoCopula)
library (Cairo)
library (extrafont)
library (ggnewscale)

source("plot_model_effect.R")
source("prepare_bird_data_with_Z_filter.R")
source("plot_alpha_diversity.R")
source("plot_coefs_with_traits.R")
source("plot_coefs_with_traits_for_publication.R")
source("multiplot.R")
# source_url("https://raw.githubusercontent.com/ronchen1/ron_functions/d2d0bab4dc65f29e53d316ae4098d0469711905b/R/multiplot.R")

# define font name
SoN_fontname <- "Almoni ML v5 AAA"

# Define the output folder path
output_path <- file.path("..","output", "batha")

# Check if the output folder exists, and create it if it doesn't
if (!dir.exists(output_path)) {
  dir.create(output_path, recursive = TRUE)
  message("Folder '", output_path, "' has been created.")
} else {
  message("Folder '", output_path, "' already exists.")
}
```

```{r load using function}
out <- prepare_bird_data()
var_names <- names(out)
for (i in 1:length(var_names)) {
  assign(var_names[i], out[[i]])
}
rm(out)
P <- P_byplot[grepl("Dwarf",unit),][order(year,site)][,site:=factor(site)][,habitat:=factor(habitat)]
P_no_rare <- P_byplot_no_rare[grepl("Dwarf",unit),][order(year,site)][,site:=factor(site)][,habitat:=factor(habitat)]
```

## Herbaceous and Dwarf-Shrub Vegetation

Factors are proximity to agriculture (plantations and orchards) and time.
Total 4 campaigns.
2 regions: herbaceous batha (grassland) and dwarf-shrub batha (low shrubland).
5 sites per region (10 sites total) with 6 plots per site (60 plots total).
**Raw data**
Total abundance: `r Araw[grepl("Dwarf",unit),sum(total_count)]`
Number of observations: `r Araw[grepl("Dwarf",unit),.N]`
Total richness: `r Araw[grepl("Dwarf",unit),uniqueN(SciName)]`

**Filtered data**
Total abundance: `r A[grepl("Dwarf",unit),sum(count_under_250)]`
Number of observations: Cannot calculate because observations were summed by plot and year.
Total richness: `r A[grepl("Dwarf",unit) & count_under_250>0,uniqueN(SciName)]`


### Model gma, abundance and richness

Richness done *with* rare species.
Abundance and mean abundance done *without* rare species.
Full models include cosine and sine of the time difference from June 21st (in radians).

#### richness

Explore data and plot mean-variance plot.
There is a strong relationship, indicating that employing GLMs is the proper way to analyze, rather than OLS (assumption of homogeneity is violated).

```{r richness data explore}
# include rare species in analysis
P.anal <- copy(P) # set a fixed variable name for analysis, if want to switch between data WITH rare species and data WITHOUT rare species then only change once here
print("RICHNESS WITH RARE SPECIES")
plot_alpha_diversity(P, x_val = "agriculture", y_val = "richness", ylab_val = "richness", xlab_val = "agriculture")

dotchart(P.anal$richness, groups = P.anal$agriculture, pch = as.numeric(P.anal$agriculture), ylab = "agriculture", xlab = "richness")
IVs <- c("richness", "year_ct","habitat","site","agriculture", "td_sc", "cos_td_rad", "sin_td_rad", "h_from_sunrise", "cos_hsun", "sin_hsun", "monitors_name",
         "wind", "precipitation", "temperature", "clouds")
kable(summary(P.anal[,..IVs]))%>% kable_styling(font_size = 11)
```

no observation for all 4 weather variables.
many NAs for sampling time of day variables. exclude from model.

```{r richness data explore 2}
pairs(P.anal[,lapply(X = .SD,FUN = as.numeric),.SDcols=IVs[1:12]])
kable(cor(P.anal[,lapply(X = .SD,FUN = as.numeric),.SDcols=IVs[1:12]], use = "pairwise.complete.obs")) %>% kable_styling(font_size = 11)
```

Fit Poisson glm, check for existence of overdispersion

```{r model specification richness, warning=FALSE}
mdl_r.poiss.int <- glm(data = P.anal, formula = richness ~ agriculture * year_ct + cos_td_rad + sin_td_rad , family = poisson)
od <- mdl_r.poiss.int$deviance/mdl_r.poiss.int$df.residual
print("Estimating overdispersion parameter phi: (Res. Dev.)/(n-p) where n=number of observations; p=number of parameters in the model.")
print(paste0('od = ',od))
```

Overdispersion parameter is <1 (underdispersion) and therefore Poisson is preferable to negative binomial.

```{r model specification richness 2}
m0 <- glm(data = P.anal, formula = richness ~ agriculture * year_ct + cos_td_rad + sin_td_rad + cos_hsun + sin_hsun, family = poisson)
summ(m0)
```

cosine and sine of hours from sunrise - remove them because of many NAs.

```{r model specification richness 3}
m0 <- glm(data = P.anal, formula = richness ~ agriculture * year_ct + cos_td_rad + sin_td_rad + habitat, family = poisson)
me0 <- glmer(data = P.anal, formula =  richness ~ agriculture * year_ct + cos_td_rad + sin_td_rad + (1|habitat), family = poisson) # model with site as random variable failed to converge
summary(me0)
```

Mixed model with site as random variable did not converge. Mixed model with habitat as random variable converged.
perform stepwise model selection of poisson mixed model.

```{r model selection richness}
drop1(me0)
```

drop cosine.

```{r}
me1 <- glmer(data = P.anal, formula =  richness ~ agriculture * year_ct + sin_td_rad + (1|habitat), family = poisson)
drop1(me1)
```

drop agriculture : year.

```{r model selection richness 3}
me1 <- glmer(data = P.anal, formula =  richness ~ agriculture + year_ct + sin_td_rad + (1|habitat), family = poisson)
drop1(me1)
```

drop year.

```{r model selection richness 4}
me1 <- glmer(data = P.anal, formula =  richness ~ agriculture  + sin_td_rad + (1|habitat), family = poisson)
drop1(me1)
```

agriculture and time of year remain.
Final model:

```{r model selection richness 5}
summary(me1)
ranef(me1)
dotplot(ranef(me1, condVar=TRUE))
plot(me1)
qqmath(me1)
# scale location plot
plot(me1, sqrt(abs(resid(.))) ~ fitted(.),
     type = c("p", "smooth"),
     main="scale-location",
     par.settings = list(plot.line =
                           list(alpha=1, col = "red",
                                lty = 1, lwd = 2)))
```

```{r model interpretation richness}
summ(me1, exp = TRUE, digits = 3)

effplot_agr <- plot_model_effect(P.anal = P.anal, m = me1, eff2plot = "agriculture", plot_points = TRUE, export_plot = TRUE, fontname = SoN_fontname, outpath = output_path)

iv_agr_abs <- diff(effplot_agr$data$richness)
iv_agr_rel <- iv_agr_abs / as.data.table(effplot_agr$data)[agriculture=="Far",richness] * 100

# m_coef <- coef(me1)
# X <- unique(data.table(samp_date = as_date(as.duration(P.anal$timediff_Jun21+172)), contrib = exp(m_coef$site[1,"cos_td_rad_c"]*P.anal$cos_td_rad_c + m_coef$site[1,"sin_td_rad_c"]*P.anal$sin_td_rad_c))[order(samp_date)][,contrib:=contrib-min(contrib)])
# plot(X$samp_date, X$contrib,xlab = "time of year", ylab = "contribution to richness", type = "b")
```

**There is a statistically significant effect for proximity to agriculture.**

Near plots have on average `r iv_agr_abs` more species than far plots, which is `r iv_agr_rel` percent higher.

**No statistically significant effect for the other factors tested.**

#### geometric mean of abundance

Explore data.
Exclude time of day because of high number of NAs.

```{r gma data explore}
# exclude rare species in analysis
P.anal <- copy(P_no_rare)

print("GEOMETRIC MEAN ABUNDANCE WITHOUT RARE SPECIES")
plot_alpha_diversity(P.anal, x_val = "agriculture", y_val = "gma", ylab_val = "gma", xlab_val = "agriculture")

dotchart(P.anal$gma, groups = P.anal$agriculture, pch = as.numeric(P.anal$agriculture), ylab = "agriculture", xlab = "gma")
IVs <- c("gma", "year_ct","habitat","site","agriculture", "td_sc", "cos_td_rad", "sin_td_rad")
kable(summary(P.anal[,..IVs]))%>% kable_styling(font_size = 11)
```

Fit glm, compare gamma, gaussian (poisson inappropriate because response is not discrete)

```{r model specification gma, warning=FALSE}
mdl_g.gamma.int <- glm(data = P.anal, formula = gma ~ agriculture * year_ct + cos_td_rad + sin_td_rad + habitat, family = Gamma)
mdl_g.gauss.int <- glm(data = P.anal, formula = gma ~ agriculture * year_ct + cos_td_rad + sin_td_rad + habitat, family = gaussian)
plot(mdl_g.gamma.int,which = 1:3,sub.caption = "gamma")
plot(mdl_g.gauss.int,which = 1:3,sub.caption = "gaussian")
```

Gamma seems better than gaussian.
exclude four outlier points: 27, 49, 69, 199
Fit fixed and mixed models.

```{r model specification gma 2}
P.anal <- P.anal[!c(27,49,69,199)]
m0 <- glm(data = P.anal, formula = gma ~ agriculture * year_ct + cos_td_rad + sin_td_rad + habitat, family = Gamma)
me0 <- glmer(data = P.anal, formula =  gma ~ agriculture * year_ct + cos_td_rad + sin_td_rad + (1|habitat), family = Gamma)
```

Mixed model with site as random variable did not converge. Mixed model with habitat as random variable converged:

```{r model specification gma 3}
summary(me0)
```

perform stepwise model selection of Gamma model.

```{r model selection gma}
drop1(me0)
```

drop cosine.

```{r}
me1 <- glmer(data = P.anal, formula =  gma ~ agriculture * year_ct + sin_td_rad + (1|habitat), family = Gamma)
drop1(me1)
```

drop agriculture : year.

```{r model selection gma 3}
me1 <- glmer(data = P.anal, formula =  gma ~ agriculture + year_ct + sin_td_rad + (1|habitat), family = Gamma)
drop1(me1)
```

drop distance from agriculture

```{r model selection gma 4}
me1 <- glmer(data = P.anal, formula =  gma ~ year_ct + sin_td_rad + (1|habitat), family = Gamma)
drop1(me1)
```

year and sampling-time-of-year remain.
This is the final model:

```{r model selection gma 5}
summary(me1)
ranef(me1)
dotplot(ranef(me1, condVar=TRUE))
plot(me1)
qqmath(me1)
# scale location plot
plot(me1, sqrt(abs(resid(.))) ~ fitted(.),
     type = c("p", "smooth"),
     main="scale-location",
     par.settings = list(plot.line =
                           list(alpha=1, col = "red",
                                lty = 1, lwd = 2)))
```


```{r model interpretation gma}
summ(me1,digits=3)

effplot_year <- plot_model_effect(P.anal = P.anal, m = me1, eff2plot = "year_ct", export_plot = TRUE, fontname = SoN_fontname, outpath = output_path)
effplot_year <- plot_model_effect(P.anal = P.anal, m = me1, eff2plot = "year_ct", plot_points = FALSE, export_plot = TRUE, fontname = SoN_fontname, outpath = output_path)

effplot_year$data$gma
2.049748/3.246487 - 1
(max(effplot_year$data$gma) - min(effplot_year$data$gma))/max(effplot_year$data$gma)

iv_year_abs <- as.data.table(effplot_year$data)[,diff(range(gma))/diff(range(year_ct))]
iv_year_rel <- as.data.table(effplot_year$data)[,exp(log(min(gma)/max(gma))/diff(range(year_ct)))]*100
```

**significant temporal decrease in gma. No effect found for distance from agriculture.**

GMA is decreasing on average in `r iv_year_abs` individuals per species per year, which is an average rate of decrease of `r 100-iv_year_rel`% per year.

#### abundance

Explore data

```{r abundance data explore}
# exclude rare species in analysis
P.anal <- copy(P_no_rare)

print("ABUNDANCE WITHOUT RARE SPECIES")
plot_alpha_diversity(P.anal, x_val = "agriculture", y_val = "abundance", ylab_val = "abundance", xlab_val = "agriculture")

dotchart(P.anal$abundance, groups = P.anal$agriculture, pch = as.numeric(P.anal$agriculture), ylab = "agriculture", xlab = "abundance")
IVs <- c("abundance", "year_ct","habitat","site","agriculture", "td_sc", "cos_td_rad", "sin_td_rad")
kable(summary(P.anal[,..IVs]))%>% kable_styling(font_size = 11)
```


```{r model specification abundance, warning=FALSE, eval=FALSE}
mdl_a.po <- glm(data = P.anal, formula = abundance ~ agriculture * year_ct + cos_td_rad + sin_td_rad + habitat, family = poisson)
od <- mdl_a.po$deviance/mdl_a.po$df.residual
print("Estimating overdispersion parameter phi: (Res. Dev.)/(n-p) where n=number of observations; p=number of parameters in the model.")
print(paste0('od = ',od))
plot(mdl_a.po,which = 1:2,sub.caption = "poisson")
```

PHI>1, hence choose negative binomial.
Fit fixed and mixed models.
Choose mixed model if possible, otherwise choose a model with fixed-effects only.

```{r model specification abundance 2}
m0 <- glm.nb(data = P.anal, formula = abundance ~ agriculture * year_ct + cos_td_rad + sin_td_rad + habitat)
me0 <- glmer.nb(data = P.anal, formula =  abundance ~ agriculture * year_ct + cos_td_rad + sin_td_rad + (1|habitat))
```

Mixed model converged:

```{r model specification abundance 3}
summary(me0)
```

Perform stepwise model selection of mixed model.

```{r model selection abundance}
drop1(me0)
```

drop cosine.

```{r}
me1 <- glmer.nb(data = P.anal, formula =  abundance ~ agriculture * year_ct + sin_td_rad + (1|habitat))
drop1(me1)
```

drop agriculture : year.

```{r}
me1 <- glmer.nb(data = P.anal, formula =  abundance ~ agriculture + year_ct + sin_td_rad + (1|habitat))
drop1(me1)
```

agriculture, year and sampling time of year remain.
The final model:

```{r model selection abundance 2}
summary(me1)
ranef(me1)
dotplot(ranef(me1, condVar=TRUE))
plot(me1)
qqmath(me1)
# scale location plot
plot(me1, sqrt(abs(resid(.))) ~ fitted(.),
     type = c("p", "smooth"),
     main="scale-location",
     par.settings = list(plot.line =
                           list(alpha=1, col = "red",
                                lty = 1, lwd = 2)))
```

Interpretation of abundance model:

```{r model interpretation abundance, warning=FALSE}
summ(me1,digits=3, exp=T)

effplot_agr <- plot_model_effect(P.anal = P.anal, m = me1, eff2plot = "agriculture", plot_points = TRUE, export_plot = TRUE, fontname = SoN_fontname, outpath = output_path)
iv_agr_abs <- diff(effplot_agr$data$abundance)
iv_agr_rel <- iv_agr_abs / as.data.table(effplot_agr$data)[agriculture=="Far",abundance] * 100

effplot_year <- plot_model_effect(P.anal = P.anal, m = me1, eff2plot = "year_ct", export_plot = TRUE, fontname = SoN_fontname, outpath = output_path)

iv_year_abs_per_year <- as.data.table(effplot_year$data)[,diff(range(abundance))/diff(range(year_ct))]
iv_year_rel_total <- (max(effplot_year$data$abundance) - min(effplot_year$data$abundance))/max(effplot_year$data$abundance)

m_coef <- fixef(me1)
X <- unique(data.table(samp_date = as_date(as.duration(P.anal$timediff_Jun21+172)), contrib = exp(m_coef["sin_td_rad"]*P.anal$sin_td_rad))[order(samp_date)])
plot(X$samp_date, X$contrib,xlab = "time of year", ylab = "Effect on abundance (multiplier)", type = "b")
```


**Significant higher abundance near agriculture, and overall decrease in total abundance**

Near plots have on average `r iv_agr_abs` more individuals than far plots, which is `r iv_agr_rel` percent higher.

Over the entire monitoring period, total abundance decreases `r iv_year_rel_total`%, which is `r 1-exp(coef(me1)$habitat[1,"year_ct"])`% per year.
On average, total abundance decreased `r iv_year_abs_per_year` individuals per year.

#### community analysis using package MVabund

```{r explore single observations}
abu_by_spp.dwarf <- abu_by_spp[grepl("Dwarf",unit),]
abu_by_spp_no_rare.dwarf <- abu_by_spp_no_rare[grepl("Dwarf",unit),]

spp <- abu_by_spp.dwarf[,.SD[,(length(col_names)+1):ncol(abu_by_spp.dwarf)]]
# filter out species with zero counts (were either not observed or are rare and were excluded)
spp <- spp[,.SD,.SDcols = colSums(spp)>0]
spp <- mvabund(spp)

spp_no_rare <- abu_by_spp_no_rare.dwarf[,.SD[,(length(col_names)+1):ncol(abu_by_spp_no_rare.dwarf)]]
# filter out species with zero counts (were either not observed or are rare and were excluded)
spp_no_rare <- spp_no_rare[,.SD,.SDcols = colSums(spp_no_rare)>0]
spp_no_rare <- mvabund(spp_no_rare)

env_data <- abu_by_spp_no_rare.dwarf[,..col_names]
env_data[,site:=factor(site)][,habitat:=factor(habitat)]

try(expr = plot(spp ~ abu_by_spp.dwarf$agriculture,overall.main="raw abundances", transformation = "no"))

plot(sort(spp_no_rare[spp_no_rare>0]))
dotchart(sort(A[grepl("Dwarf",unit) & count_under_250>=30,count_under_250]))
meanvar.plot(spp, xlab = "mean abundance of a given species across sites", ylab = "variance of the abundance of a given species across sites")
```

start model specification:

```{r community analysis MVabund model specification}
mva_m0.po <- manyglm(formula = spp_no_rare ~ agriculture * year_ct + cos_td_rad + sin_td_rad, family = "poisson", data = env_data)
mva_m0.nb <- manyglm(formula = spp_no_rare ~ agriculture * year_ct + cos_td_rad + sin_td_rad, family = "negative.binomial", data = env_data)
aic_dt <- data.table(nb = mva_m0.nb$aic, po=mva_m0.po$aic)
colMeans(aic_dt)
print("POISSON")
plot(mva_m0.po, which=1:3)
print("NEGATIVE BINOMIAL")
plot(mva_m0.nb, which=1:3)
```

negative binomial model is better than poisson according to residuals and AIC comparison.

```{r community analysis MVabund model specification 2}
mva_m0 <- mva_m0.nb
mva_m1 <- manyglm(formula = spp_no_rare ~ agriculture * year_ct + cos_td_rad + sin_td_rad +site, data = env_data)
aic_dt$nb1 <- mva_m1$aic
mva_m2 <- manyglm(formula = spp_no_rare ~ agriculture * year_ct + cos_td_rad + sin_td_rad +habitat, data = env_data)
aic_dt$nb2 <- mva_m2$aic
colMeans(aic_dt)
```

The addition of the explanatory variable 'site' is improving the AIC of the model the most.
stepwise selection of model:

```{r community analysis MVabund model selection}
drop1(mva_m1)
```

drop agriculture:year

```{r community analysis MVabund model selection 2}
mva_m1 <- manyglm(formula = spp_no_rare ~ agriculture + year_ct + cos_td_rad + sin_td_rad + site, data = env_data)
drop1(mva_m1)
```

final model includes agriculture, year, sampling time of year and site.

```{r community analysis MVabund model validation}
plot(mva_m1, which=1:3)
```

```{r community analysis MVabund model interpretation}
# summ_m1 <- summary(mva_m1, nBoot = 99)
# saveRDS(summ_m1, file = "../output/20240225_Dwarf_MVabund_summary_output_99iter.rds")
# anov_m1.uni <- anova(mva_m1, p.uni = "adjusted", nBoot = 999, show.time = "all")
# saveRDS(anov_m1.uni, file = "../output/20240225_Dwarf_MVabund_anova_output_999iter.rds")
summ_m1 <- readRDS(file = "../output/20240225_Dwarf_MVabund_summary_output_99iter.rds")
anov_m1.uni <- readRDS(file = "../output/20240225_Dwarf_MVabund_anova_output_999iter.rds")
print(summ_m1)
print(anov_m1.uni)
```

**All factors (agriculture, year, time of year and site) have a statistically significant effect on community composition.**

Below plots for crested lark, house sparrow and corn bunting

```{r community analysis MVabund interpretation galcri}
####Galerida cristata - עפרוני מצויץ####

plot_alpha_diversity(P = abu_by_spp.dwarf, x_val = "agriculture", y_val = "Galerida cristata", ylab_val = "Galerida cristata")
d.Galerida.cristata <- cbind(env_data,data.table(Galerida.cristata = spp_no_rare[,"Galerida.cristata"]))
d.Galerida.cristata <- d.Galerida.cristata[site %in% d.Galerida.cristata[,.(totabu=sum(Galerida.cristata)), by=site][totabu>0,site]]
m.Galerida.cristata <- glm.nb(formula = Galerida.cristata~agriculture + year_ct + cos_td_rad + sin_td_rad + site, data = d.Galerida.cristata)
print("Galerida cristata")
print("manyglm coefficients")
print(coef(mva_m1)[,"Galerida.cristata"])
print("species-specific glm")
print(coef(m.Galerida.cristata))

summary(m.Galerida.cristata)

#Distance from agri
Galerida_agri_plot <- effect_plot(model = m.Galerida.cristata, data=d.Galerida.cristata, pred = agriculture, interval = T, plot.points = F, jitter = c(0.1,0), point.size = 3, cat.pred.point.size = 4, partial.residuals = F, colors = "Qual1", main.title = "Temporal effect on abundance of Galerida cristata")
Galerida_agri_plot

Galerida_agri <- as.data.table(Galerida_agri_plot$data$Galerida.cristata)
 
(max(Galerida_agri)/min(Galerida_agri) - 1) * 100

#Far vs near - abundance far from agriculture is 113.65% higher compared to near
(3.834883/1.794881) - 1

#or 2.13 fold difference
3.834883/1.794881
```

```{r community analysis MVabund interpretation pasdom}
####Passer domesticus - דרור הבית####

plot_alpha_diversity(P = abu_by_spp.dwarf, x_val = "agriculture", y_val = "Passer domesticus", ylab_val = "Passer domesticus")
d.Passer.domesticus <- cbind(env_data,data.table(Passer.domesticus = spp_no_rare[,"Passer.domesticus"]))
d.Passer.domesticus <- d.Passer.domesticus[site %in% d.Passer.domesticus[,.(totabu=sum(Passer.domesticus)), by=site][totabu>0,site]]
m.Passer.domesticus <- glm.nb(formula = Passer.domesticus~agriculture + year_ct + cos_td_rad + sin_td_rad + site, data = d.Passer.domesticus)
print("Passer domesticus")
print("manyglm coefficients")
print(coef(mva_m1)[,"Passer.domesticus"])
print("species-specific glm")
print(coef(m.Passer.domesticus))

effect_plot(model = m.Passer.domesticus, data=d.Passer.domesticus, pred = year_ct, interval = T, plot.points = F, jitter = c(0.1,0), point.size = 3, cat.pred.point.size = 4, partial.residuals = F, colors = "Qual1", main.title = "Temporal effect on abundance of Passer domesticus")

summary(m.Passer.domesticus)
```

```{r community analysis MVabund interpretation embcal}
####Emberiza calandra - גבתון עפרוני####

plot_alpha_diversity(P = abu_by_spp.dwarf, x_val = "agriculture", y_val = "Emberiza calandra", ylab_val = "Emberiza calandra")
d.Emberiza.calandra <- cbind(env_data,data.table(Emberiza.calandra = spp_no_rare[,"Emberiza.calandra"]))
d.Emberiza.calandra <- d.Emberiza.calandra[site %in% d.Emberiza.calandra[,.(totabu=sum(Emberiza.calandra)), by=site][totabu>0,site]]
m.Emberiza.calandra <- glm.nb(formula = Emberiza.calandra~agriculture + year_ct + cos_td_rad + sin_td_rad + site, data = d.Emberiza.calandra)
print("Emberiza calandra")
print("manyglm coefficients")
print(coef(mva_m1)[,"Emberiza.calandra"])
print("species-specific glm")
print(coef(m.Emberiza.calandra))

print(summary(m.Emberiza.calandra))


Emberiza_year_plot <- effect_plot(model = m.Emberiza.calandra, data=d.Emberiza.calandra, pred = year_ct, interval = T, plot.points = F, jitter = c(0.1,0), point.size = 3, cat.pred.point.size = 4, partial.residuals = F, colors = "Qual1", main.title = "Temporal effect on abundance of Emberiza calandra")
Emberiza_year_plot

# Decrease of 28.54%

emberiza_time <- as.data.table(Emberiza_year_plot$data$Emberiza.calandra)
(1 - (min(emberiza_time)/max(emberiza_time))) * 100

(max(Emberiza_year_plot$data$Emberiza.calandra) - min(Emberiza_year_plot$data$Emberiza.calandra))/max(Emberiza_year_plot$data$Emberiza.calandra)
```

Corn bunting (Emberiza calandra) has nearly insignificant temporal trend in manyglm anova, and far from significant temporal trend in individual model. keep p-value because relies on many points. 2014 was sampled in July, other campaigns in april - may, hence sinus has important effect.

```{r community analysis MVabund interpretation lansen}
####lanius senator - חנקן אדום ראש####

plot_alpha_diversity(P = abu_by_spp.dwarf, x_val = "agriculture", y_val = "Lanius senator", ylab_val = "Lanius senator")
d.Lanius.senator <- cbind(env_data,data.table(Lanius.senator = spp_no_rare[,"Lanius.senator"]))
d.Lanius.senator <- d.Lanius.senator[site %in% d.Lanius.senator[,.(totabu=sum(Lanius.senator)), by=site][totabu>0,site]]
m.Lanius.senator <- glm.nb(formula = Lanius.senator~agriculture + year_ct + cos_td_rad + sin_td_rad + site, data = d.Lanius.senator)
print("Lanius senator")
print("manyglm coefficients")
print(coef(mva_m1)[,"Lanius.senator"])
print("species-specific glm")
print(coef(m.Lanius.senator))

Lanius_eff_plot <- effect_plot(model = m.Lanius.senator, data=d.Lanius.senator, pred = year_ct, interval = T, plot.points = F, jitter = c(0.1,0), point.size = 3, cat.pred.point.size = 4, partial.residuals = F, colors = "Qual1", main.title = "Temporal effect on abundance of Lanius senator")
Lanius_eff_plot

summary(m.Lanius.senator)

# Increase of 1.7%

as.data.table(Lanius_eff_plot$data)[,(max(Lanius.senator)/min(Lanius.senator))] - 1

# Lanius_eff_plot$data$Lanius.senator
(0.06873098 /0.06757608  -1) * 100


#or 1.01 fold increase

as.data.table(Lanius_eff_plot$data)[,(max(Lanius.senator)/min(Lanius.senator))]
```

Lanius senator woodchat shrike has far from significant temporal trend in individual glm and only 104 individuals observed. replace p-value with p-value from individual glm.

```{r}
anov_m1.uni$uni.p["year_ct","Lanius.senator"] <- summary(m.Lanius.senator)$coefficients["year_ct","Pr(>|z|)"] %>% round(digits = 2)
```

```{r community analysis MVabund interpretation}
####Chloris chloris - ירקון####

d.Chloris.chloris <- cbind(env_data,data.table(Chloris.chloris = spp_no_rare[,"Chloris.chloris"]))
d.Chloris.chloris <- d.Chloris.chloris[site %in% d.Chloris.chloris[,.(totabu=sum(Chloris.chloris)), by=site][totabu>0,site]]
m.Chloris.chloris <- glm.nb(formula = Chloris.chloris~agriculture + year_ct + cos_td_rad + sin_td_rad + site, data = d.Chloris.chloris)
print("Chloris chloris")
print("manyglm coefficients")
print(coef(mva_m1)[,"Chloris.chloris"])
print("species-specific glm")
print(coef(m.Chloris.chloris))


chloris_eff_plot <- effect_plot(model = m.Chloris.chloris, data=d.Chloris.chloris, pred = agriculture, interval = T, plot.points = T, jitter = c(0.1,0), point.size = 3, cat.pred.point.size = 4, partial.residuals = F, colors = "Qual1", main.title = "Spatial effect on abundance of Chloris chloris")

#Far vs near - abundance near agriculture is 580.19% higher compared to far

chloris_agri <- as.data.table(chloris_eff_plot$data$Chloris.chloris)

(max(chloris_agri)/min(chloris_agri) - 1) *100

(0.52246394/0.07681123) - 1 

#or 6.8 fold difference
0.52246394 / 0.07681123

####Pycnonotus xanthopygos - בולבול ממושקף####

d.Pycnonotus.xanthopygos <- cbind(env_data,data.table(Pycnonotus.xanthopygos = spp_no_rare[,"Pycnonotus.xanthopygos"]))
d.Pycnonotus.xanthopygos <- d.Pycnonotus.xanthopygos[site %in% d.Pycnonotus.xanthopygos[,.(totabu=sum(Pycnonotus.xanthopygos)), by=site][totabu>0,site]]
m.Pycnonotus.xanthopygos <- glm.nb(formula = Pycnonotus.xanthopygos~agriculture + year_ct + cos_td_rad + sin_td_rad + site, data = d.Pycnonotus.xanthopygos)
print("Pycnonotus xanthopygos")
print("manyglm coefficients")
print(coef(mva_m1)[,"Pycnonotus.xanthopygos"])
print("species-specific glm")
print(coef(m.Chloris.chloris))


Pycnonotus_eff_plot <- effect_plot(model = m.Pycnonotus.xanthopygos, data=d.Pycnonotus.xanthopygos, pred = agriculture, interval = T, plot.points = T, jitter = c(0.1,0), point.size = 3, cat.pred.point.size = 4, partial.residuals = F, colors = "Qual1", main.title = "Spatial effect on abundance of Pycnonotus xanthopygos")

pycno_agri <- as.data.table(Pycnonotus_eff_plot$data$Pycnonotus.xanthopygos)

#Far vs near - abundance near agriculture is 347.03% higher compared to far

(max(pycno_agri)/min(pycno_agri) - 1) * 100

Pycnonotus_eff_plot$data$Pycnonotus.xanthopygos
(0.8514907/0.1904754  ) - 1

#or 4.47 fold difference
0.8514907 / 0.1904754 

####Streptopelia.decaocto - תור צווארון####

d.Streptopelia.decaocto <- cbind(env_data,data.table(Streptopelia.decaocto = spp_no_rare[,"Streptopelia.decaocto"]))
d.Streptopelia.decaocto <- d.Streptopelia.decaocto[site %in% d.Streptopelia.decaocto[,.(totabu=sum(Streptopelia.decaocto)), by=site][totabu>0,site]]
m.Streptopelia.decaocto <- glm.nb(formula = Streptopelia.decaocto~agriculture + year_ct + cos_td_rad + sin_td_rad + site, data = d.Streptopelia.decaocto)
print("Streptopelia decaocto")
print("manyglm coefficients")
print(coef(mva_m1)[,"Streptopelia.decaocto"])
print("species-specific glm")
print(coef(m.Streptopelia.decaocto))


Streptopelia_eff_plot <- effect_plot(model = m.Streptopelia.decaocto, data=d.Streptopelia.decaocto, pred = agriculture, interval = T, plot.points = T, jitter = c(0.1,0), point.size = 3, cat.pred.point.size = 4, partial.residuals = F, colors = "Qual1", main.title = "Spatial effect on abundance of Streptopelia decaocto")

#Far vs near - abundance near agriculture is 81.93% higher compared to far

decaocto_agri <- as.data.table(Streptopelia_eff_plot$data$Streptopelia.decaocto)
(max(decaocto_agri)/min(decaocto_agri) - 1) * 100

print(Streptopelia_eff_plot$data$Streptopelia.decaocto)
(3.099531/1.703636) - 1

#or 1.81 fold difference
3.099531/1.703636

```

```{r community analysis MVabund model interpretation - plot coefficients}
coefp <- merge(data.table(t(coef(mva_m1)/log(2)),keep.rownames=TRUE), data.table(t(anov_m1.uni$uni.p), keep.rownames = TRUE), by = "rn") # coefficients are on log2 scale to simplify interpretation, yet keep presenetation on log scale rather than linear. The link function used for poisson is log -> above 0 is a positive trend and below 0 is negative trend

# add total species abundance
coefp <- merge(coefp,as.data.table(colSums(spp_no_rare),keep.rownames = TRUE), by.x = "rn", by.y = "V1")

# add hebrew name and traits
sci_heb <- unique(A[,.(SciName,HebName)])
sci_heb[,SciName:=make.names(SciName)]
coefp <- merge(coefp, sci_heb, by.x = "rn", by.y = "SciName")

# set column names based on the predictor variables
colnames(coefp)[grepl("rn",colnames(coefp))] <- "SciName"
colnames(coefp)[grepl("Intercept).x",colnames(coefp))] <- "intercept.coef"
colnames(coefp)[grepl("Intercept).y",colnames(coefp))] <- "intercept.p"
colnames(coefp)[colnames(coefp)=="settlementsNear"] <- "settlementsNear.coef"
colnames(coefp)[colnames(coefp)=="settlements"] <- "settlements.p"
colnames(coefp)[colnames(coefp)=="agricultureNear"] <- "agricultureNear.coef"
colnames(coefp)[colnames(coefp)=="agriculture"] <- "agriculture.p"
colnames(coefp)[grepl("cos_td_rad.x",colnames(coefp))] <- "cos_td_rad.coef"
colnames(coefp)[colnames(coefp)=="cos_td_rad.y"] <- "cos_td_rad.p"
colnames(coefp)[grepl("sin_td_rad.x",colnames(coefp))] <- "sin_td_rad.coef"
colnames(coefp)[grepl("sin_td_rad.y",colnames(coefp))] <- "sin_td_rad.p"
colnames(coefp)[colnames(coefp)=="year_ct.x"] <- "year_ct.coef"
colnames(coefp)[colnames(coefp)=="year_ct.y"] <- "year_ct.p"
colnames(coefp)[colnames(coefp)=="V2"] <- "species_abundance"
colnames(coefp)[colnames(coefp)=="dunesshifting"] <- "dunesShifting.coef"
colnames(coefp)[colnames(coefp)=="dunes"] <- "dunes.p"
colnames(coefp)[colnames(coefp)=="site"] <- "site.p"
colnames(coefp)[colnames(coefp)=="subunitCarmel"] <- "subunitCarmel.coef"
colnames(coefp)[colnames(coefp)=="subunitGalilee"] <- "subunitGalilee.coef"
colnames(coefp)[colnames(coefp)=="subunit"] <- "subunit.p"
colnames(coefp)[colnames(coefp)=="habitatWadi"] <- "habitatWadi.coef"
colnames(coefp)[colnames(coefp)=="habitat"] <- "habitat.p"

plot_coefs_agr <- plot_coefs_with_traits(D = A_no_rare[grepl("Dwarf",unit)],
                                         cp = coefp[,.(SciName,agricultureNear.coef,agriculture.p,species_abundance)],
                                         plot_title = "Effect of proximity to agriculture on species abundance",
                                         plot_xlabel = expression(paste("coefficient of NEAR agriculture, (log"[2]," scale)")),
                                         mark_batha = TRUE)
plot_coefs_agr

p_agr <- plot_coefs_with_traits_for_publication(D = A_no_rare[grepl("Dwarf",unit)],
                                                cp = coefp[,.(SciName,agricultureNear.coef,agriculture.p,species_abundance)],
                                                plot_title = "Effect of proximity to agriculture on species abundance",
                                                plot_xlabel = expression(paste("coefficient of NEAR agriculture, (log"[2]," scale)")),
                                                mark_batha = TRUE,
                                                show_legend = FALSE,
                                                export_plot = TRUE,
                                                fontname = SoN_fontname,
                                                effect_str = "agriculture",
                                                outpath = output_path)
p_agr

plot_coefs_year <- plot_coefs_with_traits(D = A_no_rare[grepl("Dwarf",unit)],
                                          cp = coefp[,.(SciName,year_ct.coef,year_ct.p,species_abundance)],
                                          plot_title = "Temporal effect on species abundance",
                                          plot_xlabel = expression(paste("coefficient of year counter, log"[2]," scale)")),
                                          mark_batha = TRUE)
plot_coefs_year

p_year <- plot_coefs_with_traits_for_publication(D = A_no_rare[grepl("Dwarf",unit)],
                                                 cp = coefp[,.(SciName,year_ct.coef,year_ct.p,species_abundance)],
                                                 plot_title = "Temporal effect on species abundance",
                                                 plot_xlabel = expression(paste("coefficient of year counter, log"[2]," scale)")),
                                                 mark_batha = TRUE,
                                                 show_legend = FALSE,
                                                 export_plot = TRUE,
                                                 fontname = SoN_fontname,
                                                 effect_str = "year",
                                                 outpath = output_path)
p_year
```

**Significant effect of plantation agriculture proximity: synanthrope / invasive species near agriculture.**
**Most species showing negative coefs, consistent with overall decrease in gma and total abundance.**

Significant negative temporal effect on corn bunting and lesser whitethroat (but see species-specific plots above). Positive significant trend for woodchat shrike, but the magnitude is very small (also see above).

Worth noting that the following have consistent results across several units:

crested lark, negative, in Sfar and Batha.
house sparrow, negative, in Sfar, Batha and Maquis.
red-rumped swallow, positive, in sfar, batha and maquis.
palestine sunbird, negative, in sfar, batha and maquis.
goldfinch, negative, in maquis and batha.

```{r export manyglm p-values}
write.csv(coefp[,.(SciName,HebName,year_ct.p, agriculture.p)], file = "../output/batha/species_composition_p_values.csv", row.names = F, fileEncoding = "MS-HEBR")
```